---
execute:
  echo: false
---

# Results

TODO: I've provided all the comparison results, most of which are not
statistically significant. Maybe I should also provide grand mean, summary
results.

```{r include=FALSE}
load("./fuel_data_modeling.rda")
load("./sprout_modeling.rda")
load("./regen_visualize.rda")

library(tidyverse)
library(emmeans)
library(ggdist)
library(purrr)
library(stringr)
library(tibble)
library(dplyr)
library(ggplot2)
library(tibble)
library(patchwork)
library(tinytable)
```

```{r}

# Given a glmmTMB model and one of "cond", "disp", or "zi", print model formula
# optionally including labels and link functions
format_model_formula <- function(mod, component, label = FALSE) {
  comp_label <- switch(component,
    cond = c("Conditional:", ""),
    disp = c("Dispersion:", "(log)"),
    zi = c("Hurdle:", "(logit)")
  )
  form <- format(formula(mod, component = component)) |>
    paste0(collapse = "") |>
    str_replace("\\s{2,}", " ")
  if (label) return(paste(comp_label[1], form, comp_label[2]))
  form
}

# given a glmmTMB model and extra component model types, print conditional and
# component model formula and link functions
print_model_info <- function(mod, components = c("disp", "zi")) {
  comp <- c("cond", components)
  fam <- paste0(
    "Family: ", mod$modelInfo$family$family,
    " (", mod$modelInfo$family$link, ")"
  )
  cat(fam, "\n")
  walk(comp, \(x) cat(format_model_formula(mod, x, label = TRUE), "\n"))
}

treat_labels <- c(gs = "GS", ld = "LD", ha = "HA", hd = "HD")
species_labels <- c(LIDE = "Tanoak", SESE = "Redwood", other = "Other", df =
  "Douglas-fir", rw = "Redwood", to = "Tanoak")
year_labels <- c(`1` = "Year 1", `5` = "Year 5", `10` = "Year 10")
plot_labels <- labeller(
  spp = species_labels,
  year =  year_labels,
  treat = treat_labels
)

standard_model_col_names <- c(
  LCL = "lower.CL", UCL = "upper.CL", LCL = "asymp.LCL", UCL = "asymp.UCL",
  emmean = "response", emmean = "the.emmean", treat = "treatment"
)

# produce comparisons of emmeans grid and convert to a data frame to plot the
# results
emm_to_df <- function(emm) {
  plot(emm, comparisons = TRUE, plotit = FALSE) |>
  as_tibble() |>
  rename(any_of(standard_model_col_names))
}

# Common mapping for all plots, allows for specification of x-axis variable
emm_mapping <- function(x = treat, group = NULL, color = NULL) {
  aes({{x}}, emmean, ymin = LCL, ymax = UCL, xend = {{x}}, group = {{group}},
  color = {{color}})
}

# Common elements of all plots: comparison arrows, and CI's 
draw_emm <- function(data = ., position = "identity", show_legend = FALSE) {
  list(
    geom_linerange(
      color = "gray70",
      linewidth = 2.3,
      show.legend = show_legend,
      position = position
    ),
    geom_segment(
      aes(yend = rcmpl),
      color = "#4444d1",
      lineend = "butt",
      linejoin = "bevel",
      arrow = arrow(length = grid::unit(2.5, "mm")),
      linewidth = 1,
      na.rm = TRUE,
      position = position
    ),
    geom_segment(
      aes(yend = lcmpl),
      color = "#4444d1",
      lineend = "butt",
      linejoin = "bevel",
      arrow = arrow(length = grid::unit(2.5, "mm")),
      linewidth = 1,
      na.rm = TRUE,
      position = position
    ),
    geom_point(
      color = "white",
      shape = "circle filled",
      size = 2,
      fill = "black",
      position = position
    )
  )
}

# Common elements of all plots: comparison arrows, and CI's 
draw_emm2 <- function(data = ., position = position_dodge(width = .7)) {
  list(
    geom_linerange(
      linewidth = 2.3,
      position = position
    ),
    geom_point(
      color = "white",
      shape = "circle filled",
      size = 2,
      fill = "black",
      position = position
    )
  )
}


```

## Regeneration composition

### Basal area

Composition of regeneration in terms of total basal area of each species was
modeled as a gamma distribution with a log link with fixed effects for
treatment, species, and their interaction and random effects for sites and
macro-plots. Dispersion was modeled separately as a function of species, using a
log link and the rate of zeros was modeled using the logit link, for each
species (@lst-regen-ba).

::: {#lst-regen-ba}
```{r}
print_model_info(mba)
```
::: 

According to this model, we would expect five times as much basal area of other
species in LD compared to HA (p = 0.034), and six times more other species
basal area in HD compared to HA (p = 0.014), but the presence of other species
was generally less than 0.75 m^2^ ha^-1^.

We expect 5 times more redwood regeneration basal area in the GS treatment
compared to HA (p = 0.03), and 7 times compared to HD (p = 0.02).

For tanoak, we expect about twice as much regeneration basal area in the GS
treatment compared to HA (p = 0.03).

```{r}
#| label: fig-regen-ba
#| fig-cap: >
#|   Basal area (m^2^ ha^-1^) modeled at the vegetation plot level for four
#|   harvest treatments and four species classes (n = 16). Gray bars represent
#|   the 95% confidence interval, black dots---the mean, and non-overlapping
#|   blue arrows signify statistical significance (α = 0.05).

emmeans(mba, "treat", by = "spp", type = "response") |> # pairs()
 emm_to_df() |>
 ggplot(emm_mapping()) +
 facet_wrap(~ spp, ncol = 2, labeller = plot_labels, scales = "free") +
 draw_emm() +
 scale_x_discrete(labels = toupper) +
 labs(x = "Treatment", y = expression(Basal ~ Area ~ (m^2%.%ha^-1)))

```

@fig-regen-ba-rw-to shows the same model as @fig-regen-ba, but with an emphasis
on treatment comparisons between redwood and tanoak. This shows that we expect
on average, five times greater redwood basal area than tanoak basal area in the
GS treatment (p < 0.001), and about 2.7 times in the LD treatment (p = 0.039).
The two species were most similar in the HD treatment, where redwood basal area
is expected to be 1.4 times that of tanoak (p = 0.56). Redwood variability,
indicated by the size of 95% confidence intervals, is much greater than that of
tanoak's across treatments, and redwood variability was greater in GS and LD
treatments than in the HA and HD treatments (@fig-regen-ba).

```{r}
#| label: fig-regen-ba-rw-to
#| fig-cap: >
#|   Basal area (m^2^ ha^-1^) modeled at the vegetation plot level for four
#|   harvest treatments and two species classes (n = 16). Gray bars represent
#|   the 95% confidence interval, black dots---the mean, and non-overlapping
#|   blue arrows signify statistical significance (α = 0.05).

emmeans(
  mba,
  "spp",
  by = "treat",
  at = list(spp = c("rw", "to")),
  type = "response"
) |> # pairs()
  emm_to_df() |>
  ggplot(emm_mapping(x = spp)) +
  facet_wrap(~ treat, ncol = 2, labeller = plot_labels, scales = "free") +
  draw_emm() +
  scale_x_discrete(labels = species_labels) +
  labs(x = "Species", y = expression(Basal ~ Area ~ (m^2%.%ha^-1)))

```

### Other species

Other speceis included grand fir, madrone, and California wax-myrtle, of which
there was a total of 23, 28, and 16 observations across our 16 macro plots
(comprising 64 tree density plots). Generally, each plot had between 0 and 9
observations of other species, except for one macro plot with the LD treatment,
which had 16 observations (data not shown).

### Douglas-fir counts

Counts of regenerating Douglas-fir seedlings per vegetation plot (n = 16) were
analyzed for differences between harvest treatments using a negative binomial
response with a log link, fixed effects for treatment, random effects for site
and macro-plot (@lst-df-counts).

::: {#lst-df-counts}
```{r}
print_model_info(mdf, NULL)
```
:::

This model for Douglas-fir counts does not result in any statistically
significant differences between treatments. Generally, we expect about 2
seedlings per 4-meter-radius plot, or about 413 seedlings per hectare
(@fig-df-counts).

```{r}
#| label: fig-df-counts
#| fig-cap: >
#|   Vegetation plot level counts of regenerating Douglas-fir seedlings in four
#|   harvest treatments 10 years after harvest (n = 16). Results have been
#|   scaled to stems per hectare (4-meter radium plots).

# predict(mdf_sel) |> mean() |> exp() * per_ha["corner"]

emmeans(mdf, "treat", type = "response") |>
 emm_to_df() |>
 mutate(across(where(is.double), \(x) x * per_ha["corner"])) |>
 ggplot(emm_mapping()) +
 draw_emm() +
 scale_x_discrete(labels = toupper) +
 labs(x = "Treatment", y = expression(Stems%.%ha^-1))

```

## Sprout heights

### Height increment

The selected height increment model used a normal response distribution on the
identity link. It included treatment, growth period, species, and the
interaction of species and growth period as fixed effects. A random intercept
was included for tree (multiple observations) and macro-plot, and a random slope
was included for species. The dispersion parameter for the response was modeled
(with a log link) as a function of treatment, growth period, species and all
three-way interactions (@lst-sprout-ht-inc).  

::: {#lst-sprout-ht-inc}
```{r}
print_model_info(minc_sel, "disp")
```
:::

The model selected based on AIC lacks a treatment x species interaction,
suggesting that there is not evidence that treatments affected species
differentially. It also lacks a treatment x year interaction. This means that
there was not enough evidence to support that treatment was related to changes
in growth rate.

The presence of treatment in the model (0.001 ≤ p < 0.03) suggests that
the levels of treatment were associated with different growth rates across
species and years. And the species x year interaction (p < 0.001) suggests
changes in growth rates are different for redwood and tanoak
(@fig-sprout-ht-inc-treat).

Averaging over growth periods, treatment specific height increments for redwood
ranged from 0.66 to 0.86 m yr^-1^, and for tanoak, from 0.29 to 0.49, with the
slowest growth in the HD treatment and the fastest in the GS treatment. Height
increment was greater in the GS treatment than the HA and HD treatments by about
0.19 m yr^-1^ (p < 0.001). The LD treatment was intermediate, but not
statistically distinguishable from the other treatments (0.13 < p < 0.28).

```{r}
#| label: fig-sprout-ht-inc-treat
#| fig-cap: >
#|   Estimated marginal means for the effect of harvest treatment on redwood and
#|   tanoak sprout height increment, averaged over two growth periods, ten years
#|   after harvest. Gray bars represent confidence intervals and statistical
#|   significance (α = 0.05) is indicated by non-overlapping blue arrows.

emmeans(minc_sel, "treat", by = "spp", type = "response") |> # pairs()
  emm_to_df() |>
  ggplot(emm_mapping()) +
  draw_emm() +
  facet_wrap(~ spp, labeller = plot_labels) +
  labs(x = "Treatment", y = expression(Height~Increment~(m%.%yr^-1)))

```

Redwood growth slowed from 0.80 to 0.67 m yr^-1^ in the second period and tanoak
slowed from 0.39  to 0.34 m yr^-1^.

Redwood grew faster than tanoak, but slowed down more relative to it in the
second period. Height increment for redwood was 0.42 m yr^-1^ greater in the
first period and 0.33 m yr^-1^ greater in the second period than tanoak height
increment (@fig-sprout-ht-inc-year).


```{r}
#| label: fig-sprout-ht-inc-year
#| fig-cap: >
#|   Estimated marginal means for the effect of growth period on redwood and
#|   tanoak sprout height increment, averaged over four harvest treatments, from
#|   years 1 to 5, and years 5 to 10 after harvest, plotted alongside actual
#|   data. Gray bars represent confidence intervals and statistical significance
#|   (α = 0.05) is indicated by non-overlapping blue arrows.

emmeans(minc_sel, c("year"), by = "spp", type = "response") |>
  emm_to_df() |>
  ggplot(emm_mapping(x = year)) +
  draw_emm(position = position_nudge(x = -0.15)) +
  geom_dots(data = dinc, aes(year, ht_inc, ymin = NULL, ymax = NULL), alpha = 0.4) +
  facet_wrap(~ spp, labeller = plot_labels) +
  scale_x_discrete(labels = c(`5` = "1 - 5", `10` = "5 - 10")) +
  labs(
    x = "Growth Period (years)",
    y = expression(Height ~ increment ~ (m %.% year^-1))
  )

```

### Height at year 10

Sprout heights at year 10 were modeled with a normal response and a log link.
The best model included species and treatment, but no interactions in the fixed
effects. This suggests that treatments do not affect species differentially. It
also included a model for dispersion (log link) that had species, treatment, and
their interaction as predictors (@lst-sprout-ht-yr-10).

::: {#lst-sprout-ht-yr-10}
```{r}
print_model_info(ht10_sel, "disp")
```
:::

Because the best model did not contain a species x treatment interaction,
comparisons between treatments is the same for both species. The GS treatment
resulted in greater heights in year 10 than the other treatments (0.001 < p <
0.04). Predicted mean height for redwood ranged from 10.29 m in the GS treatment
to 6.16 m in the HD treatment. For tanoak, predicted mean height ranged from
5.12 in the GS treatment to 3.04 in the HD treatment. Predicted mean heights
followed the pattern GS > LD > HA > HD (@fig-sprout-ht-yr-10).

```{r}
#| label: fig-sprout-ht-yr-10
#| fig-cap: >
#|   Predicted mean height and 95% confidence intervals (gray bars) for redwood
#|   and tanoak stump sprouts 10 years after harvest using four different
#|   harvest treatments. Non-overlapping blue arrows indicate statistically
#|   significant differences between treatments within a species.

emmeans(ht10_sel, "treat", by = "spp", type = "response") |>
  emm_to_df() |>
  ggplot(emm_mapping()) +
  draw_emm(position = position_nudge(x = -0.07)) +
  geom_dots(
    data = filter(dht, year == 10),
    aes(x = treat, y = ht),
    inherit.aes = FALSE
  ) +
  facet_wrap(~ spp, labeller = plot_labels) +
  labs(x = "Treatment", y = "Height (m)")

```

## Fuels

### Pre-pct

Gamma distributed, linear multi-level models, with a log link were used for all
six fuel class responses. Random intercepts were specified for three levels of
nesting, representing sites, treatment blocks, and transect corners. All models
except for the duff & litter model included a hurdle model to account for
zero, which was modeled with a logit link. For the 10-hr fuel model, the hurdle
portion was modeled as a function of treatment, and for the others, it was
modeled as a single rate for all observations. The 10-hr fuel model also
included a dispersion model, which was modeled with a log link, using treatment
as a predictor (@tbl-fuel-pre-pct).

```{r}
#| label: tbl-fuel-pre-pct
#| tbl-cap: Model specifications for six fuel classes before pct.

fuel_tmb$pre |>
  transmute(
    class = fuel_class_labels[as.character(class)],
    Family = mod$modelInfo$family$family,
    Link = mod$modelInfo$family$link,
    Conditional = format_model_formula(mod, "cond"),
    "Dispersion (log)" = format_model_formula(mod, "disp"),
    "Hurdle (logit)" = format_model_formula(mod, "zi")
  ) |>
  tinytable::tt()

```

For Duff & Litter, the largest difference was between the HD and HA treatments.
The HD treatment had about 1.4 times more duff and litter (p = 0.07). Generally,
all treatments were similar, with estimated loading of around 50 Mg ha-1.
One-hour fuels were around 50% higher in the HA treatment compared to the LD and
GS treatments (p = 0.07, and p = 0.01, respectively), with mean differences of
around 0.5 Mg ha-1. Ten, hundred and thousand-hour fuels were statistically,
very similar across treatments (p = 0.7 — p = 1). Point estimates varied by
about 1, 3, and <20 Mg ha-1 for ten, hundred, and thousand-hour fuels,
respectively. Vegetative fuel loading was greatest in the GS treatment, with an
expected value of 28.5 Mg ha-1, which was about 2.7 times greater than in HA (p
= 0.01) (@fig-fuel-pre-pct).

```{r}
#| label: fig-fuel-pre-pct
#| fig-width: 8
#| fig-cap: >
#|   Estimated marginal means (black dots) confidence intervals (gray bands) and
#|   comparisons (blue arrows) of fuel loading across four treatments for six
#|   different fuel-class models. Non-overlapping blue arrows indicates
#|   statistical significance at the α = 0.05 level.

fuel_tmb$pre |>
  mutate(class = fuel_class_labels[as.character(class)]) |>
  group_map(function(x, ...) {
    emm_to_df(x$.emmeans[[1]]) |>
    ggplot(emm_mapping()) +
    draw_emm() +
    scale_x_discrete(labels = toupper) +
    labs(title = x$class, y = expression(Load~(Mg%.%ha^-1)), x = "Treatment")
  }) |>
  patchwork::wrap_plots() + patchwork::plot_layout(axes = "collect")

# inspect means and differences for vegetation fuels
# fuel_tmb$pre$pairs[[6]]
# fuel_tmb$pre$emmeans[[6]]
# fuel_tmb$pre |> filter(class == "dufflitter") |> pluck("emmeans", 1)

```

Here are the grand means, pre-PCT for each fuel type:

```{r}

grand_mean <- fuel_tmb$pre |>
  mutate(grand_mean = as_tibble(emmeans(.emmeans, ~1))) |>
  select(c(class, grand_mean)) |>
  unnest(grand_mean)
grand_mean |> tinytable::tt()

```


### Post-pct

The response for all six, post-pct fuel classes were modeled with a gamma
distribution and a log link, and included the same multi-level random effects as
for the pre-pct models. Dispersion models with treatment as the only predictor
were included for 1-hr and 100-hr fuel classes. All models included a hurdle
portion to model zeros using a logit link. For 100-hr fuels, this model included
treatment and site as predictors, and for the rest, a constant rate for all
observations was used (@tbl-fuel-post-pct).

```{r}
#| label: tbl-fuel-post-pct
#| tbl-cap: Model specifications for six fuel classes after pct.
fuel_tmb$post |>
  transmute(
    class = fuel_class_labels[as.character(class)],
    Family = mod$modelInfo$family$family,
    Link = mod$modelInfo$family$link,
    Conditional = format_model_formula(mod, "cond"),
    "Dispersion (log)" = format_model_formula(mod, "disp"),
    "Hurdle (logit)" = format_model_formula(mod, "zi")
  ) |> 
  tinytable::tt()
```

Post-pct resulted in greater stratification of treatments (@fig-fuel-post-pct).
One-hour fuels for most treatments were around 2.4 Mg ha^-1^, but the HA
treatment had around half of that amount (p = 0.01 to p = 0.02). The GS
treatment had the greatest 10-hr fuel loading with 8.8 Mg ha^-1^, which was
about 1.6, 2.3 and 2.9 times greater than the LD, HA, and HD treatments
respectively (p = 0.03, p < 0.001, for the others, respectively). The LD
treatment also had about 1.7 times more 10-hr fuels that the HD treatment (5.4
vs. 3 Mg ha^-1^, p = 0.001). Hundred-hour fuels were also greatest in the GS
treatment, with an average of about 19 Mg ha^-1^, which was about 2.6 times
greater than in the HD treatment (7 Mg ha^-1^, p < 0.001). Thousand-hour fuels
were greatest in the HD treatment, with 80 Mg ha^-1^, which was about 2.7 times
greater than the LD and HD treatments (p = 0.03 and p = 0.05, respectively).
Fuel loading for live vegetation was similar across treatments at around 2.5 Mg
ha^-1^. The pre-post vegetation difference was greatest in the GS treatment at
about 31 Mg ha^-1^, which was 2.5 and 2.8 times the HD and HA treatments,
respectively (p ≈ 0.01).

```{r}
#| label: fig-fuel-post-pct
#| fig-width: 8
#| fig-cap: >
#|   Estimated marginal means (black dots) confidence intervals (gray bars) and
#|   comparisons (blue arrows) of fuel loading across four treatments for six
#|   different fuel-class models. Non-overlapping red arrows indicates
#|   statistical significance at the α = 0.05 level. Vegetation difference
#|   equals the transect level difference in vegetation load in the pre and
#|   post-pct conditions. This represents slash fuels recruited to the forest
#|   floor following the pre-commercial thinning.

fuel_tmb$post |>
  mutate(class = fuel_class_labels[as.character(class)]) |>
  group_map(function(x, ...) {
    emm_to_df(x$.emmeans[[1]]) |>
    ggplot(emm_mapping()) +
    draw_emm() +
    scale_x_discrete(labels = toupper) +
    labs(title = x$class, y = expression(Load~(Mg%.%ha^-1)), x = "Treatment")
  }) |>
  patchwork::wrap_plots() + patchwork::plot_layout(axes = "collect")

```

Here are the grand means post-PCT for each fuel type:

```{r}

grand_mean <- fuel_tmb$pre |>
  mutate(grand_mean = as_tibble(emmeans(.emmeans, ~1))) |>
  select(c(class, grand_mean)) |>
  unnest(grand_mean)
grand_mean |> tinytable::tt()

```

Pre-commercial thinning led to a small increase in average 100-hr fuel loading,
only for the GS treatment, increased 10-hr fuels in the GS and LD treatments,
and increased 1-hr fuels for all but the HA treatment
(@fig-fuel-pct-comparison), although these results are not statistically
comparable, due to slightly different model structures.

```{r}
#| label: fig-fuel-pct-comparison
#| fig-width: 8
#| fig-cap: >
#|   Estimated marginal means (black dots) and confidence intervals (colored
#|   bars) of fuel loading across four treatments and five different fuel
#|   classes, before and after PCT. Pre- and Post PCT models within a treatment
#|   are from similar, but not necisarily identical models.

# pre and post together
fuel_tmb$pre |>
  # only use fuel classes that are the same between PCT phase
  filter(class %in% c("onehr", "tenhr", "hundhr", "thoushr", "veg")) |>
  # mutate(class = fuel_class_labels[as.character(class)]) |>
  group_map(function(x, ...) {
    # for each row in pre-treatment, get the pre-treatment data and the
    # corresponding data for post-treatment
    bind_rows(
      pre = emm_to_df(x$.emmeans[[1]]),
      post = emm_to_df(
        pluck(filter(fuel_tmb$post, class == x$class), ".emmeans", 1)
      ),
      .id = "phase"
    ) |>
    ggplot(emm_mapping(
      color = factor(phase, levels = c("pre", "post")),
      group = factor(phase, levels = c("pre", "post"))
    )) +
    draw_emm2() +
    scale_x_discrete(labels = toupper) +
    scale_color_manual(
      labels = c("Pre-PCT", "Post-PCT"),
      values = c("#d17844", "#4444d1")
    ) +
    labs(
      title = fuel_class_labels[as.character(x$class)],
      y = expression(Load~(Mg%.%ha^-1)),
      x = "Treatment",
      color = "Phase"
    )
  }) |>
  patchwork::wrap_plots() +
    guide_area() +
    patchwork::plot_layout(
      axes = "collect_y",
      axis_title = "collect",
      guides = "collect"
    )

```


