---
title: Summarize raw fuel data
format:
  html:
    code-fold: true
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(message = FALSE)
```

```{r}
#| warning: false
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(forcats)
source("./scripts/process_datasheets.r")
source("./scripts/test_funs.r")
```

Before I convert our data to fuel loads, I would like to have at least a summary
of the actual data that we collected in the field.


## Load data from datasheets

Data in quasi-csv format, described in @sec-data-entry. I'm ommitting the medium
density dispersed plot (MD), which only exists at one site.

Counts data is straight forward

Sampling cylinder data uses a mix of NA and 0 if there was no vegetation. I'm
making the height 0 if there was no vegetation. Also, if there is no vegetation,
species is "-".

I'm converting to duff/litter depth and percent litter to duff depth and litter
depth.

```{r}

transectid <- c("phase", "site", "treatment", "corner", "azi")

d <- combine_fuels_datasheets("../data") |>
  map(\(x) filter(x, treatment != "md"))

warn_duplicates(d$transects, phase, site, treatment, corner, azi)

counts <- d$transect |>
  select(all_of(transectid), slope, matches("(one|ten|hund)hr_count")) |>
  replace_na(list(slope = 0))

stations <- d$transect |>
  select(all_of(transectid), matches("\\w+[12]$")) |>
  pivot_longer(
    !all_of(transectid),
    names_to = c(".value", "station"),
    names_pattern = "(\\w+)([12])"
  ) |>
  # there was some inconsistency in whether heights were zero or blank if no veg
  # was present, here I sort that out.
  mutate(
    avg_w_ht = if_else(live_woody == 0 & dead_woody == 0, 0, avg_w_ht),
    avg_h_ht = if_else(live_herb == 0 & dead_herb == 0, 0, avg_h_ht),
    species = tolower(str_replace(species, "hb", "eh")),
    species = if_else(avg_w_ht == 0, "-", species),
    litter_depth = duff_litter * pct_litter / 100,
    duff_depth = duff_litter - litter_depth
  ) |>
  select(-c(metermark, duff_litter, pct_litter)) |>
  relocate(fbd, .after = last_col())

most_freq_string <- function(x) names(which.max(table(x)))

cwd <- d$coarse_woody
```

Heights and decay class observations are conditional on percent cover and
diameter respectively, it might make more sense to report weighted averages
here, but I'm not doing that.

For coarse woody debris, counts are averaged differently than diameter and decay
class. I'll report average diameter of *observed* particles only, and average of
counts at the transect level (including zeros). I need to ensure I'm making zero
counts explicit, as they are not in the data.

```{r}

sum_funcs <- list(
  mean = function(x) mean(x, na.rm = TRUE),
  se = function(x) mean(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
)

stations_sum <- stations |>
  group_by(phase, treatment) |>
  summarise(across(c(where(is.double), -azi), sum_funcs)) |>
  pivot_longer(
    -c(treatment, phase),
    names_to = c("class", ".value"),
    names_pattern = "(.*)_(mean|se)"
  )

counts_sum <- counts |>
  group_by(phase, treatment) |>
  summarise(across(c(where(is.double), -azi), sum_funcs)) |>
  pivot_longer(
    -c(treatment, phase),
    names_to = c("class", ".value"),
    names_pattern = "(.*)_(mean|se)"
  )

# I need to know all possible transects to know where there are zeros for coarse
# woody
all_transects <- d$transects |>
  expand(nesting(phase, site, treatment, corner, azi))

# Average decay and size across all measured logs, average counts for all
# transects
cwd_sum <- cwd |>
  group_by(phase, treatment) |>
  mutate(across(c(dia, decay), sum_funcs)) |>
  group_by(phase, treatment, site, corner, azi) |>
  summarize(count = n(), across(matches("_(se|mean)$"), first)) |>
  right_join(all_transects) |>
  replace_na(list(count = 0)) |>
  group_by(phase, treatment) |>
  summarize(across(count, sum_funcs), across(matches("_(se|mean)$"), first)) |>
  pivot_longer(
    -c(treatment, phase),
    names_to = c("class", ".value"),
    names_pattern = "(.*)_(mean|se)"
  )
```

Here is a summary of all the data we collected.

```{r}
#| label: tbl-raw-sampling-cylinders
#| tbl-cap: >
#|   Raw raw summary of data collected within sampling cylinders for each
#|   treatment (n = 64). Average heights are for live and dead vegetation
#|   particles, and don't represent an average height across an area. They are
#|   meant to be multiplied by percent cover, but are presented indepenently
#|   here to document our actual measurements.

# this pivots wider and formats mean and se to print
stations_sum |>
  mutate(across(c(mean, se), \(x) format(round(x, 1)))) |>
  mutate(mean = glue::glue("{mean} ({se})", .trim = FALSE), .keep = "unused") |>
  pivot_wider(names_from = "class", values_from = "mean") |>
  mutate(
    treatment = fct_relevel(treatment, c("gs", "ld", "ha", "hd")),
    phase = factor(phase, levels = c("prepct", "postpct"))
  ) |>
  arrange(phase, treatment) |>
  rename(
    "Phase" = phase,
    "Treatment" = treatment,
    "Live Woody Cover (%)" = live_woody,
    "Dead Woody Cover (%)" = dead_woody,
    "Average Woody Height (m)" = avg_w_ht,
    "Live Herb Cover (%)" = live_herb,
    "Dead Herb Cover (%)" = dead_herb,
    "Average Herb Height (m)" = avg_h_ht,
    "Litter Depth (cm)" = litter_depth,
    "Duff Depth (cm)" = duff_depth,
    "Fuel Bed Depth (cm)" = fbd
  ) |>
  knitr::kable(align = c("l", "l", rep("r", 9)))
```

```{r}
#| label: tbl-raw-fwd-transects
#| tbl-cap: >
#|   Raw raw summary of data collected along fuel transects for each
#|   treatment (n = 32). Transect lengths for 1-hr, 10-hr, and 100-hr fuels were
#|   1, 2, and 4 m, respectively. Redwood leaf sprays were only counted if they
#|   were ⪆2 mm.

counts_sum |>
  mutate(across(c(mean, se), \(x) format(round(x, 1)))) |>
  mutate(mean = glue::glue("{mean} ({se})", .trim = FALSE), .keep = "unused") |>
  pivot_wider(names_from = "class", values_from = "mean") |>
  mutate(
    treatment = fct_relevel(treatment, c("gs", "ld", "ha", "hd")),
    phase = factor(phase, levels = c("prepct", "postpct"))
  ) |>
  arrange(phase, treatment) |>
  rename(
    "Phase" = phase,
    "Treatment" = treatment,
    "Slope" = slope,
    "1-hr Count" = onehr_count,
    "10-hr Count" = tenhr_count,
    "100-hr Count" = hundhr_count
  ) |>
  knitr::kable(align = c("l", "l", rep("r", 4)))
```


```{r}
#| label: tbl-raw-cwd-transects
#| tbl-cap: >
#|   Raw raw summary of data collected for coarse woody debris (> 7.62 cm) along
#|   10-m transects for each treatment (n = 32). Diameter and decay class are
#|   for observed particles averaged at the treatment level (48 ≤ n ≤ 84).
#|   Counts are made at the transect level and averaged at the treatment level
#|   (n = 32).

cwd_sum |>
  mutate(across(c(mean, se), \(x) format(round(x, 1)))) |>
  mutate(mean = glue::glue("{mean} ({se})", .trim = FALSE), .keep = "unused") |>
  pivot_wider(names_from = "class", values_from = "mean") |>
  mutate(
    treatment = fct_relevel(treatment, c("gs", "ld", "ha", "hd")),
    phase = factor(phase, levels = c("prepct", "postpct"))
  ) |>
  arrange(phase, treatment) |>
  rename(
    "Phase" = phase,
    "Treatment" = treatment,
    "Count" = count,
    "Diameter (cm)" = dia,
    "Decay Class" = decay
  ) |>
  knitr::kable(align = c("l", "l", rep("r", 3)))
```

```{r}
save(d, file = "summarize_collected_data.rda")
```

