---
title: Reneration data summary
format:
  html:
    code-fold: true
---

```{r}
#| include: false
source("./scripts/utils.r")
```

```{r}
#| output: false

library(tidyr)
library(tidyverse)
library(ggdist)
library(patchwork)
library(vegan)
library(glmmTMB)
library(DHARMa)

load("regen_wrangled.rda")
```

```{r include=FALSE}

species_labels <- c(to = "Tanoak", rw = "Redwood", df = "Douglas-fir", 
  other = "Other")

plot_labels <- labeller(
  spp = species_labels
)

```

Some functions and constants that I need later on, these might should get moved
to their own function and sourced.

```{r}

# The metric foresters constant: diameter in cm to area of a circle in square
# meters
for_const <- function(dbh) {
  pi * dbh^2 / (4 * 10000)
}

# I need a mapping so that each species gets a unique color, I'll make this a
# function that takes a palette function from the scales package and returns a
# named vector

spp_col <- function(data, .palfun, ...) {
  pal <- .palfun(...)
  spp <- unique(data$spp)
  purrr::set_names(pal(length(spp)), spp)
}


```

## Objectives

- Determine if there are significant differences in species composition, or size
  class distirbution between treatments

This objective is somewhat vaugue because I am not familiar with the statistical
techniques used for the weighted (diameter), multi-variate (species)
distribution data that we have.

For starters, I will look at summaries of size class distributions by species. I
will report densities in terms of trees per hectare (tph). Our plots were fixed
area, 4-m-radius plots, or 50.27 m^2^. Each stem in a veg. plot represents
198.94 stems/ha. As shown in @fig-sampling-design, there are:

- 64 veg. plots in the experiment
- 16 in a treatment
- 4 in a macro plot

This plot is a little difficult to interpret beccause of the large difference in
counts from the smallest to largest stems. It might make more sense to look at
regen smaller than 5 cm separately.

```{r}
#| label: fig-raw-hist-regen-1
#| fig-cap: >
#|   Histogram of regen by species and size class for each treatment.

# expansion factors at different aggregation levels
per_ha <- c(
  all = 1/64, treat = 1/16, plot = 1/4, corner = 1) * (10000 / (16 * pi)
)

p <- regen |>
  ggplot(aes(dbh, fill = spp)) +
  stat_bin(aes(y = after_stat(count) * per_ha["treat"]), bins = 10) +
  facet_wrap(vars(treat), nrow = 1) +
  labs(y = expr(trees%.%ha^-1))
p

```

Splitting the analysis at 5-cm-diameter stems shows a clear trend in treatments
from most to least available light. Over this gradient, the number of 1.3-cm
stems increases, and the 3.8-cm stems decrease, almost proportionally --- and
most of this is due to tanoak. For >= 5-cm stems, the effect is a fairly
homogenous decrease in the number of stems. There are few tanoak, Douglas-fir
are only in the brightest areas. Mostly redwood (number of stems) shows a strong
correlation with available light.

```{r}
#| label: fig-raw-hist-regen-2
#| fig-cap: >
#|   Histograms of regen by species and size class. On top are stems less than
#|   five centimeters in diameter, and stems greater than five centimeters are
#|   on the bottom.

raw_hist2 <- function(group, bin_breaks, scale_breaks, text_pos) {
  generate_facet_totals <- function(data) {
    summarize(data,
      .by = treat,
      n = paste("Total: ", round(n() * per_ha["treat"])),
    )
  }
  colors <- spp_col(regen, scales::brewer_pal, palette = "Set2")
  regen |>
    filter({{ group }}) |>
    ggplot(aes(dbh, fill = forcats::fct_reorder(spp, spp, length))) +
    stat_bin(aes(y = after_stat(count) * per_ha["treat"]), breaks = bin_breaks) +
    ggpp::geom_text_npc(
      data = generate_facet_totals,
      aes(npcx = "center", npcy = "top", label = n, fill = NULL)
    ) +
    facet_wrap(vars(treat), nrow = 1) +
    labs(y = expression(trees %.% ha^-1), x = "dbh (cm)", fill = "spp") +
    scale_x_continuous(breaks = scale_breaks) +
    scale_y_continuous(expand = expansion(c(0.05, 0.2))) +
    scale_fill_manual(values = colors) +
    theme_bw()
}

raw_hist2(dbh < 5, seq(0, 17.5, 2.5), seq(0, 15, 5), c(2.5, 6000)) /
raw_hist2(dbh >= 5, seq(0, 17.5, 2.5), seq(0, 15, 5), c(27, 600))

```

This is interesting, but I wonder what this would look like in terms of basal
area, which is more biologically relevant than number of stems.

```{r}

local({
  bin_breaks <- seq(0, 17.5, 2.5)
  scale_breaks <- seq(0, 15, 5)
  facet_total_label <- function(data) {
    summarize(data,
      .by = treat,
      ba_ha = round(sum(per_ha["treat"] * for_const(dbh)), 1)
    )
  }
  regen |>
    ggplot(aes(dbh, fill = forcats::fct_reorder(spp, dbh, .fun = ~sum(.x^2)))) +
    stat_bin(aes(weight = per_ha["treat"] * for_const(dbh)), breaks = bin_breaks) +
    ggpp::geom_text_npc(
      data = facet_total_label,
      mapping = aes(npcx = "center", npcy = "top", label = ba_ha, fill = NULL)
    ) +
    facet_wrap(vars(treat), nrow = 1) +
    labs(y = expr(m^2 %.% ha^-1), x = "dbh (cm)", fill = "spp") +
    scale_y_continuous(expand = expansion(c(0.05, 0.2))) +
    scale_x_continuous(breaks = scale_breaks) +
    scale_fill_manual(
      values = spp_col(regen, scales::brewer_pal, palette = "Set2")
    ) +
    theme_bw()
})

```

# Community analysis: Species and abundance (BA)

What I would like to know is, are there significant differences between
treatments in terms of number of stems, or basal area each species and size
class. For instance, a hypothetical result might be: 

*GS has 30% more redwood stems and 20% more redwood basal area than LD*

*The HD treatment resulted in more Douglas-fir than the LD and GS treatments*

Above I visualize basal area by treatment, species, and size class. Here I'll
simplify and visualize just treatment and species.

```{r}
#| label: fig-raw-abundance-ba

# I'm expanding my data to ensure that any plot missing a species observation is
# recorded as zeros.
colors <- spp_col(regen, scales::brewer_pal, palette = "Set2")

regen |>
  summarise(
    .by = c(site, treat, plot, spp),
    ba_ha = round(sum(per_ha["corner"] * for_const(dbh)), 2)
  ) |>
  complete(nesting(site, treat, plot), spp, fill = list(ba_ha = 0)) |>
  ggplot(aes(treat, ba_ha, fill = spp)) +
  geom_boxplot() +
  facet_wrap(~ spp, scales = "free_y") +
  scale_fill_manual(values = colors)

```

Here is another plot to get a better sense of what the data within treatment blocks
looks like

```{r}

regen |>
  summarise(
    .by = c(site, treat, plot, spp),
    ba_ha = round(sum(per_ha["corner"] * for_const(dbh)), 2)
  ) |>
  complete(nesting(site, treat, plot), spp, fill = list(ba_ha = 0)) |>
  ggplot(aes(treat, ba_ha, fill = spp)) +
  geom_boxplot() +
  facet_wrap(~ spp + site, scales = "free_y") +
  scale_fill_manual(values = colors)
```

## Permanova

While it's not possible to include fully nested random effects using the
`adonis2` function, it seems important to try to include at least some
accounting for grouping. I think `site` might be the most beneficial grouping
level to capture because it captures differences in aspect, and also contains
more observations than lower grouping levels.

I will start out by visualizing the differences associated with sites.

This first plot shows that the whiskey spring site has generally higher basal
area and that the waldo n site has an especially high basal area for the just
the ld treatment.

```{r}
#| fig-width: 9
#| fig-height: 5

total_ba_ha <- regen |>
  summarize(.by = c(site, treat, plot),
    ba_ha = round(sum(per_ha["corner"] * for_const(dbh)), 2)
  )

dotplot(
  ba_ha ~ treat,
  total_ba_ha,
  groups = site,
  type = c("p", "a"),
  xlab = "Treatment",
  ylab = expression(m^2%.%ha^-1),
  auto.key = list(columns = 2, lines = TRUE),
  jitter.x = TRUE
)

```

This figure shows that the site:treatment interaction is likely significant.
There are similar trends across sites, but their magnitude is different
depending on treatment.

```{r}
#| fig-width: 9
#| fig-height: 5

dotplot(
  ba_ha ~ site,
  total_ba_ha,
  groups = treat,
  type = c("p", "a"),
  xlab = "Site",
  ylab = expression(m^2%.%ha^-1),
  auto.key = list(columns = 2, lines = TRUE),
  jitter.x = TRUE
)

```

Next, I'll prepare the species data for analysis

```{r}

ba_ha <- regen |>
  summarise(
    .by = c(site, treat, plot, spp),
    ba_ha = round(sum(per_ha["corner"] * for_const(dbh)), 2)
  ) |>
  pivot_wider(names_from = spp, values_from = ba_ha) |>
  mutate(across(where(is.double), ~ replace_na(.x, 0)))

```

I took these from the help for `adonis2`, I'm not quite sure what aspect of the
data they are showing yet.

```{r}

mod <- metaMDS(ba_ha[c("rw", "to", "df", "other")], trace = FALSE)
plot(mod)

ordiellipse(mod, ba_ha$site, kind = "ehull", label = TRUE)
ordispider(mod, ba_ha$site, lty = 3, col = "red")

```

In the following models, the strata argument is used to: "Constrain perumtations
within groups." The first two models don't include the site as strata. 

When I include site as a fixed effect, p-values increase slightly.

```{r}

perms <- 500

adonis2(
  ba_ha[c("rw", "to", "df", "other")] ~ treat,
  data = ba_ha,
  permutations = perms
)

adonis2(
  ba_ha[c("rw", "to", "df", "other")] ~ site + treat,
  data = ba_ha,
  permutations = perms
)

```

If I put site in `strata`, I get similar results.

```{r}
adonis2(
  ba_ha[c("rw", "to", "df", "other")] ~ treat,
  data = ba_ha,
  strata = ba_ha$site,
  permutations = perms
)

```

If I include the interaction of site:treatment as a fixed effect, it is not
significant.

```{r}

with(ba_ha, adonis2(
  cbind(rw, to, df, other) ~ site + site:treat + treat,
  permutations = perms
))


```

I think we can safely conclude that ther are some significant differences in
basal are composition (I assume at the alpha = 0.05 level) between treatments.

Next, we might consider where these differences (in abundance defined as basal
area) are. 

# Species abundance model

We could ask about:

- Differences between species within a treatment or
- Differences between treatment within a species. 

I will be focussing on the latter which is in keeping with the theme of the rest
of my analysis. These differences have already been visualized above in
@fig-raw-abundance-ba.

Again, I'm making zero observations explicit, effectively making our data
**presence-absence**, rather than presence only.

```{r}

regen_pa <- regen |>
  summarize(.by = c(site, treat, plot, spp),
    ba_ha = round(sum(per_ha["corner"] * for_const(dbh)), 2)
  ) |>
  complete(nesting(site, treat, plot), spp, fill = list(ba_ha = 0))

regen_po <- regen |>
  summarize(.by = c(site, treat, plot, spp),
    ba_ha = round(sum(per_ha["corner"] * for_const(dbh)), 2)
  )

```

So, perhaps with our 16 observations per species/treatment, we can detect some
differences in basal area with a multi-level model that accounts for
non-independence of our sampling design.

```{r}

library(emmeans)
m1 <- glmmTMB(ba_ha ~ treat * spp + (1 | site/treat/plot), regen_pa)
res1 <- simulateResiduals(m1)
plot(res1)

m2 <- glmmTMB(
  ba_ha ~ treat * spp + (1 | site/treat),
  regen_pa,
  dispformula = ~ spp
)
res2 <- simulateResiduals(m2)
plot(res2)

m3 <- glmmTMB(
  ba_ha ~ treat * spp + (1 | site/treat),
  regen_pa,
  family = ziGamma(link = "log"),
  dispformula = ~ spp,
  ziformula = ~ spp
)
plot(simulateResiduals(m3))

m4 <- glmmTMB(
  ba_ha ~ treat * spp + (1 | site/treat),
  regen_pa,
  family = ziGamma(link = "log"),
  dispformula = ~ spp + treat,
  ziformula = ~ spp
)

res4 <- simulateResiduals(m4)
plot(res4)
plotResiduals(res4, form = regen_pa$spp)
plotResiduals(res4, form = regen_pa$treat)
plotResiduals(res4, form = interaction(regen_pa$treat, regen_pa$spp))

AIC(m1, m2, m3, m4)

```
**Select this model**

```{r}
mba <- m3
```

Model three is the simplest, has the lowest AIC and has well-behaved residuals.

Here I plot some simulations from this model to see if it generates plausible
data.

```{r}

regen_pa |>
  mutate(simulate(mba, nsim = 5)) |>
  rename(obs = ba_ha) |>
  pivot_longer(where(is.double), names_to = "type", values_to = "ba_ha") |>
  ggplot(aes(
    treat, ba_ha, # fill = type
    group = interaction(type, treat),
    fill = forcats::fct_collapse(type, sim = paste0("sim_", 1:5))
  )) +
  geom_boxplot() +
  facet_wrap(~ spp, scales = "free_y") +
  labs(fill = "obs/sim")

```

Finally, significant differences?

We'd expect 5 times as much basal area of other species in LD compared to HA, and
interestingly, 6 times more other species basal area in HD compared to HA.

We expect 5 times more redwood regen basal area in GS compared to HA, and 7
times compared to HD.

For tanoak, we expect about twice as much regeneration basal area in the GS treatment compared to HA.

```{r}
#| label: fig-species-ba-model
#| fig-cap: 

emmeans(mba, c("spp", "treat"), by = "spp", type = "response") |>
  plot(comparisons = TRUE, plotit = FALSE) |> as_tibble() |>
  ggplot(aes(treat, the.emmean)) +
  facet_wrap(~ spp, scales = "free", labeller = plot_labels) +
  geom_linerange(
    aes(ymin = asymp.LCL, ymax = asymp.UCL),
    color = "gray70",
    linewidth = 2.3,
    show.legend = FALSE
  ) +
  geom_segment(
    aes(xend = treat, yend = rcmpl),
    color = "#4444d1",
    lineend = "butt",
    linejoin = "bevel",
    arrow = arrow(length = grid::unit(2.5, "mm")),
    linewidth = 1
  ) +
  geom_segment(
    aes(xend = treat, yend = lcmpl),
    color = "#4444d1",
    lineend = "butt",
    linejoin = "bevel",
    arrow = arrow(length = grid::unit(2.5, "mm")),
    linewidth = 1
  ) +
  geom_point(color = "black", size = 2) +
  labs(x = "Treatment", y = expression(Basal ~ Area ~ (m^2%.%ha^-1))) +
  scale_x_discrete(labels = toupper)

```

```{r}

emmeans(mba, c("spp", "treat"), by = "spp", type = "response") |>
contrast("pairwise") |>
as_tibble() |>
filter(p.value < 0.05) |>
relocate(spp)
```

## Douglas-fir counts

For Douglas-fir in particular, we are interested in the counts of stems more
than the basal area. This, requires another model. A counts model.

Heres the raw counts. There doesn't appear to be any treatment differences in
the number of Douglas-fir seedlings per plot.

```{r}
#| label: fig-raw-dougfir-counts
#| fig-cap: >
#|   Summary of raw count data for Douglas-fir expressed in stems per
#|   hectare, for Douglas-fir seedlings in four harvest treatments 10 years
#|   after harvest.

regen_df <- regen |>
  count(site, treat, plot, spp) |>
  complete(nesting(site, treat, plot), spp, fill = list(n = 0)) |>
  filter(spp == "df")

regen_df |>
  ggplot(aes(treat, n * per_ha["corner"])) +
  geom_boxplot() +
  labs(x = "Treatment", y = expression(Stems%.%ha^-1))

```

A negative binomial model seems to fit acceptably when modeling raw counts (not
TPH). I was not able to achive model convergence with the specification of zero
inflation or dispersion formulae.

```{r}

m1 <- glmmTMB(
  n ~ treat + (1 | site/treat),
  regen_df,
  family = nbinom1()
)

res1 <- simulateResiduals(m1)
plot(res1)

```

**Select this model**

```{r}
mdf <- m1
```

There are no statistically detectable differences

```{r}

emmeans(m1, "treat")

```

The model for Douglas-fir seedling counts does not result in any statistically
significant differences between treatments. Generally, we expect about six
seedlings per 4-meter-radius plot, or 1,000 to 1,400 seedling per hectare.

```{r}
#| label: tbl-dougfir-counts-p-values
#| tbl-cap: >
#|   P-values for pairwise contrasts of Douglas-fir seedling counts in
#|   in four different harvest treatments 10 years after harvest.

emmeans(m1, "treat") |>
  contrast("pairwise") |>
  as_tibble() |>
  tinytable::tt()

```

Save data from this page for future use

```{r}

save(regen, spp_col, per_ha, for_const, mba, regen_pa, mdf, regen_df, file = "regen_visualize.rda")

```
