---
title: Sprout data exploration
format:
  html:
    code-fold: true
---

::: {.content-visible when-format="html"}

First, we'll load some libraries and our data.

```{r}
#| output: false

library(ggdist)
library(ggridges)
library(tidyverse)
library(patchwork)

load("wrangled_sprouts.rda")
source("./scripts/rnd_color_brewer.r")

```

```{r}
#| include: false

ggplot2::theme_update(
  panel.grid.major = ggplot2::element_blank(),
  panel.grid.minor = ggplot2::element_blank()
)

```

## Objectives

1. For each species/treatment combination, is there a difference in height
   increment between the first and second periods?
2. Is there a difference in redwood or tanoak sprout heights at year 10, if so,
   what is the magnitude of the difference?

## Visualize data

### Tree heights

The following figures reveal possible trends in the raw data.

#### Species/treatment/year

@fig-data-summary-treatments reveals fine, species specific differences between
treatments and general trends over time. It shows that over time, LD and GS
treatments have more taller trees than the other treatments, with GS having the
most. This is true for both redwood and tanoak. Additionally, for redwood it
appears that HA may have taller trees than HD, at least in year 10.

@fig-data-summary-spp shows the same information as before, but is arguably 
easier to look at. There is a general increasing trend in heights with
treatments as follows:

    HD < HA < LD < GS

Redwood is consistently taller than tanoak, and the GS treatment confers the
greatest advantage to redwood.

Across all treatments, it is also interesting to note that over times, the
height distributions, especially for redwood, seem to becoming more multi-modal
and more widely distributed. This could be due to site or plot effects, or to
microsite (within plot) effects, but it is not immediately clear why this
diverging performance should be so apparent with redwood and not tanoak. 


```{r}
#| label: fig-data-summary-treatments
#| fig-height: 8
#| fig-width: 8
#| fig-cap: >
#|   Height distributions for tanoak and redwood over time for each of four
#|   treatments. Treatments GS and LD have higher proportions of taller trees.
#|   Years refers to number of years after treatment.
#| message: false

sprouts |>
  lengthen_data("ht") |>
  ggplot(aes(ht, treat, color = treat, fill = treat)) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  geom_dots(binwidth = unit(0.019, "npc")) +
  facet_grid(vars(spp), vars(year), scales = "free", labeller = label_both) +
  labs(x = "Height (m)", y = "Treatment") +
  scale_color_brewer(palette = "Set2", aesthetics = c("color", "fill")) +
  theme(legend.position = "bottom")


```

```{r}
#| label: fig-data-summary-spp
#| fig-height: 8
#| fig-cap: >
#|   Similar to @fig-data-summary-treatments, but with an emphasis on
#|   differences between speceis responses across treatments. Comparisons are
#|   made for each year. The GS treatment appears to favor redwood response the
#|   most, but all treatments show redwoods are taller than tanoak. Years refers
#|   to number of years after treatment.
#| message: false

sprouts |>
  lengthen_data("ht") |>
  ggplot(aes(
    x = ht,
    y = fct_relevel(treat, treat_order),
    color = factor(spp, levels = c("SESE", "LIDE"))
  )) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  geom_density_ridges(alpha = .3, size = 1) +
  facet_grid(year ~ ., labeller = label_both) +
  labs(x = "Height (m)", y = "Treatment") +
  scale_color_brewer(palette = "Set2", name = "Species")
```

#### Species/year/SDI
@fig-data-summary-sdi shows that above around 400 SDI, tree heights level off.
It also implies a steady decrease in height from 0 to around 400 SDI. The
strength of the relationship appears to be increasing over time, particularly
for redwood.

```{r}
#| label: fig-data-summary-sdi
#| fig-cap: SDI vs HT for both species, across measurement years.
#| fig-width: 8.5

sprouts |>
  lengthen_data("ht") |>
  ggplot(aes(sdi_init, ht, color = factor(year))) +
  geom_point(alpha = 0.5) +
  geom_smooth(span = 1.2) +
  facet_wrap(~spp) +
  scale_color_brewer(palette = "Set2")
```

#### Species/site/plots
Lets see what the variability among sites and plots looks like, I'll focus on
year 10 only.

@fig-data-summary-sites reveals some differences between sites, particularly for
redwoods. Waldo North tends to have larger redwoods and Camp 6 has a large
proportion of smaller redwoods.

@fig-data-summary-plots shows that much of the difference in sites has to do
with one outlier plot within a site, and less about general site trends. We
should expect plots to capture a portion of the variance. Most notable is the
plot level difference between redwood and tanoak. For redwood, the large ammount
of within plot variability combined with the between plot variability obscures
the treatment (and site) effect. If you squint, there appears to be a similar
overall pattern between redwood and tanoak repsonse to treatment, but it appears
they respond differentially to certain plots.

```{r}
#| label: fig-data-summary-sites
#| fig-cap: >
#|   Distribution of heights at year 10 at each site for two species. Slabs are
#|   normalized by sample size to reflect the raw data, plotted as dots below.


sprouts |>
  ggplot(aes(ht10, site, fill = spp)) +
  stat_slab(aes(thickness = after_stat(pdf * n)), alpha = 0.6, scale = 0.7) +
  stat_dotsinterval(side = "bottom", scale = 0.7, slab_color = NA) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "year 10 height (m)", y = NULL, fill = "species")

```

```{r}
#| label: fig-data-summary-plots
#| fig-height: 7
#| fig-cap: >
#|   Same as for @fig-data-summary-sites, but for each plot (Site/treatment
#|   interaction). Grouped by treatment.

sprouts |>
  mutate(treat = fct_relevel(treat, treat_order)) |>
  ggplot(aes(
    ht10,
    site,
    fill = treat,
  )) +
  stat_slab(aes(thickness = after_stat(pdf * n)), alpha = 0.5, scale = 0.7) +
  stat_pointinterval(position = position_dodge(width = 0.5, preserve = "single")) +
  scale_color_brewer(palette = "Set2") +
  facet_wrap(~spp) +
  theme(legend.position = "bottom") +
  labs(y = NULL, x = "year 10 height (m)", fill = "treatment")

```

### Height increments

Height increments contain similar information as heights, but allow us to
compare directly between years.

@fig-data-summary-ht-inc-year shows that across treatments and species, height
growth slows down in the second period (years 5-10). This is more true for
redwood but it starts with more rapid growth than tanoak. In the most crowded
treatment (HD), redwoods height increment has become slower than tanoaks in the
second period. Also, in the second period, the high density, aggregated
treatment appears to have slightly higher (or equal) average growth increment,
which is not completely expected.

```{r}
#| label: fig-data-summary-ht-inc-year
#| fig-height: 7
#| fig-cap: >
#|   Variations in annual height growth increment between the first and second
#|   measurement periods (years 1-5, and 5-10, respectively).

sprouts |>
  lengthen_data("ht_inc") |>
  ggplot(
    aes(
      ht_inc,
      fct_relevel(treat, treat_order),
      fill = factor(year),
      color = factor(year)
    )
  ) +
  stat_slab(alpha = 0.5) +
  stat_pointinterval(
    position = position_dodge(width = 0.4, preserve = "single")
  ) +
  facet_wrap(~spp) +
  theme(legend.position = "bottom") +
  labs(
    fill = "Year",
    color = "Year",
    y = "Treatment",
    x = expression(Height~increment~(m~yr^-1))
  ) +
  scale_color_manual(
    values = rnd_color_brewer("Set2", c(1,4)),
    aesthetics = c("color", "fill")
  )

```

:::
