---
title: Sprout data exploration
format:
  html:
    code-fold: true
---

::: {.content-visible when-format="html"}

First, we'll load some libraries and our data.

```{r}
#| output: false

library(ggdist)
library(ggridges)
library(tidyverse)
library(patchwork)

load("wrangled_sprouts.rda")
source("./scripts/rnd_color_brewer.r")

```

```{r}
#| include: false

# ggplot2::theme_update(
#   panel.grid.major = ggplot2::element_blank(),
#   panel.grid.minor = ggplot2::element_blank()
# )

```

## Objectives

1. For each species/treatment combination, is there a difference in height
   increment between the first and second periods?
2. Is there a treatment difference in redwood or tanoak sprout heights at year
   10, if so, what is the magnitude of the difference?

## Visualize data

### Tree heights

The following figures reveal possible trends in the raw data.

#### Species/treatment/year


Redwood is consistently taller than tanoak, and the GS treatment confers the
greatest advantage to redwood.

Across all treatments, it is also interesting to note that over times, the
height distributions, especially for redwood, seem to becoming more multi-modal
and more widely distributed. This could be due to site or plot effects, or to
microsite (within plot) effects, but it is not immediately clear why this
diverging performance should be so apparent with redwood and not tanoak. 


```{r}
#| label: fig-data-summary-treatments
#| fig-height: 8
#| fig-width: 8
#| fig-cap: >
#|   Raw sprout heights for redwood and tanoak in four different harvest
#|   treatments, at three measurement periods over 10 years. Note the difference
#|   in x-axis scales.
#|
#| message: false

species_labels <- c(LIDE = "Tanoak", SESE = "Redwood")
year_labels <- c(`1` = "Year 1", `5` = "Year 5", `10` = "Year 10")
plot_labels <- labeller(
  spp = species_labels,
  year =  year_labels
)


sprouts |>
  lengthen_data("ht") |>
  ggplot(aes(ht, treat, color = treat, fill = treat)) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  geom_dots(binwidth = unit(0.019, "npc")) +
  facet_grid(vars(spp), vars(year), scales = "free_x", labeller = plot_labels) +
  labs(x = "Height (m)", y = "Treatment") +
  scale_color_brewer(palette = "Set2", aesthetics = c("color", "fill")) +
  theme(legend.position = "bottom")


```

This plot shows the raw data collected for ~100 redwood and tanoak sprout clumps
in each of four harvest treatments. There is a clear increasing trend with year.
The height distribution of redwood is much more dispersed than tanoak. The
dispersion of both species increases with year. Redwood are consistently taller.
There appears to be an increasing trend in height in treatments where GS > LD >
HA > HD.

```{r}
#| label: fig-data-summary-spp
#| fig-height: 8
#| fig-cap: >
#|   Density plots of raw sprout heights for redwood and tanoak in four
#|   different harvest treatments, at three measurement periods over 10 years.
#| message: false

sprouts |>
  lengthen_data("ht") |>
  ggplot(aes(
    x = ht,
    y = fct_relevel(treat, treat_order),
    color = factor(spp, levels = c("SESE", "LIDE"))
  )) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  geom_density_ridges(alpha = .3, size = 1) +
  facet_grid(year ~ ., labeller = plot_labels) +
  labs(x = "Height (m)", y = "Treatment") +
  scale_color_brewer(palette = "Set2", name = "Species", labels = species_labels)

```

#### Species/year/SDI
@fig-data-summary-sdi shows that above around 400 SDI, tree heights level off.
It also implies a steady decrease in height from 0 to around 400 SDI. The
strength of the relationship appears to be increasing over time, particularly
for redwood.

```{r}
#| label: fig-data-summary-sdi
#| fig-cap: SDI vs HT for both species, across measurement years.
#| fig-width: 8.5

sprouts |>
  lengthen_data("ht") |>
  ggplot(aes(sdi_init, ht, color = factor(year))) +
  geom_point(alpha = 0.5) +
  geom_smooth(span = 1.2) +
  facet_wrap(~spp) +
  scale_color_brewer(palette = "Set2")
```

#### Species/site/plots
Lets see what the variability among sites and plots looks like, I'll focus on
year 10 only.

@fig-data-summary-sites reveals some differences between sites, particularly for
redwoods. Waldo North tends to have larger redwoods and Camp 6 has a large
proportion of smaller redwoods.

@fig-data-summary-plots shows that much of the difference in sites has to do
with one outlier plot within a site, and less about general site trends. We
should expect plots to capture a portion of the variance. Most notable is the
plot level difference between redwood and tanoak. For redwood, the large ammount
of within plot variability combined with the between plot variability obscures
the treatment (and site) effect. If you squint, there appears to be a similar
overall pattern between redwood and tanoak repsonse to treatment, but it appears
they respond differentially to certain plots.

```{r}
#| label: fig-data-summary-sites
#| fig-cap: >
#|   Distribution of heights at year 10 at each site for two species. Slabs are
#|   normalized by sample size to reflect the raw data, plotted as dots below.


sprouts |>
  ggplot(aes(ht10, site, fill = spp)) +
  stat_slab(aes(thickness = after_stat(pdf * n)), alpha = 0.6, scale = 0.7) +
  stat_dotsinterval(side = "bottom", scale = 0.7, slab_color = NA) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "year 10 height (m)", y = NULL, fill = "species")

```

```{r}
#| label: fig-data-summary-plots
#| fig-height: 7
#| fig-cap: >
#|   Same as for @fig-data-summary-sites, but for each plot (Site/treatment
#|   interaction). Grouped by treatment.

sprouts |>
  mutate(treat = fct_relevel(treat, treat_order)) |>
  ggplot(aes(
    ht10,
    site,
    fill = treat,
  )) +
  stat_slab(aes(thickness = after_stat(pdf * n)), alpha = 0.5, scale = 0.7) +
  stat_pointinterval(position = position_dodge(width = 0.5, preserve = "single")) +
  scale_color_brewer(palette = "Set2") +
  facet_wrap(~spp) +
  theme(legend.position = "bottom") +
  labs(y = NULL, x = "year 10 height (m)", fill = "treatment")

```

### Height increments

Height increments contain similar information as heights, but allow us to
compare directly between years.

@fig-data-summary-ht-inc-year shows that across treatments and species, height
growth slows down in the second period (years 5-10). This is more true for
redwood but it starts with more rapid growth than tanoak. In the most crowded
treatment (HD), redwoods height increment has become slower than tanoaks in the
second period. Also, in the second period, the high density, aggregated
treatment appears to have slightly higher (or equal) average growth increment,
which is not completely expected.

```{r}
#| label: fig-data-summary-ht-inc-year
#| fig-height: 7
#| fig-cap: >
#|   Variations in annual height growth increment between the first and second
#|   measurement periods (years 1-5, and 5-10, respectively).

sprouts |>
  lengthen_data("ht_inc") |>
  ggplot(
    aes(
      ht_inc,
      fct_relevel(treat, treat_order),
      fill = factor(year),
      color = factor(year)
    )
  ) +
  stat_slab(alpha = 0.5) +
  stat_pointinterval(
    position = position_dodge(width = 0.4, preserve = "single")
  ) +
  facet_wrap(~spp) +
  theme(legend.position = "bottom") +
  labs(
    fill = "Year",
    color = "Year",
    y = "Treatment",
    x = expression(Height~increment~(m~yr^-1))
  ) +
  scale_color_manual(
    values = rnd_color_brewer("Set2", c(1,4)),
    aesthetics = c("color", "fill")
  )

```

:::
