---
title: Calculate fuel loading
---

We used the line intercept method for tallying fine and coarse woody debris
[@vanwagnerPracticalAspectsLine1982; @brownHandbookInventoryingDowned1974]. In
essence, with this technique we are recording the diameters of all dead, downed
woody pieces that cross an imaginary sampling plane of a specified length and
height. These diameters are convered to crossectional areas which can then be
extended to a volume by multiplying by a width. Finally, volume can be converted
to mass using specific gravity. This method requires the assumptions that a
particle is (1) cylidrical, (2) horizontal, and (3) perpendicular to the
sampling plane. While (1) may be a reasonable assumption for many woody
particles, 2 and 3 are less likely. For 2, we use a factor which allows us to
trade out this assumption, for the assumption that pieces are randomly oriented
in the horizontal plane. For 3, we use a correction factor ($\sec \theta$) based
estimates of average paticle inclination ($\theta$).

In practice, the diameter of each piece is not measured, but tallied by size
class (up to a certain size, after which it is prudent to measure individual
pieces). We used the common time-lag size classes < 0.635 cm; 0.635 - 2.54 cm;
and 2.54 - 7.62 cm. All pieces larger than 7.62 cm were measured to the nearest
cm. This requires estimating an average diameter (quadratic mean diameter is
used) for pieces in each category which can be done using emperical data
gathered from specific forest types [@brownHandbookInventoryingDowned1974], or
estimated from the sampling distribution of the tallies themselves
[@vanwagnerPracticalAspectsLine1982]. 

The resulting equation for estimating fuel load (Mg) is: 

$$Load = \frac{\sum d^2  G  k  a  c}{l}$$

where d is diameter (cm), G is specific gravity, k is a constant which accounts
for random horizontal particle orientation, the density of water and unit
conversion factors, a is the particle inclination correction factor, c is the
transect slope correction factor and l is the transect length.

The "composite" values for G (for both sound and rotten wood), a, and the
quadratic mean diameter of a size class (used for d) where provided by
@brownHandbookInventoryingDowned1974 for use when site specific data are
unknown. The values given a pertain to "non-slash". Somewhat larger values are
used for slash fuel beds.

```{r}
#| code-fold: true

params <- dplyr::tribble(
  ~source,        ~parameter, ~onehr, ~tenhr, ~hundhr, ~thoushr_s, ~thoushr_r,
  "brown",        "d2",       0.10,   1.86,   17.81,   NA,         NA,
  "brown",        "G",        0.48,   0.48,   0.40,    0.40,       0.30,
  "brown",        "a",        1.13,   1.13,   1.13,    1.00,       1.00,
  "vanwagtendonk","d2",       0.12,   1.28,   14.52,   NA,         NA,
  "vanwagtendonk","G",        0.58,   0.57,   0.53,    0.47,       0.36,
  "vanwagtendonk","a",        1.03,   1.02,   1.02,    1.02,       1.02,
)

glebocki <- dplyr::tribble(
  ~id, ~parameter, ~onehr, ~tenhr, ~hundhr, ~thoushr_s, ~thoushr_r,
  "85-0405", "a", 1.03, 1.03, 1.02, 1.01, 1.00, 
  "85-0405", "d2", 0.13, 1.32, 16.13,  NA, NA,
  "85-0405", "G", 0.45, 0.46, 0.42, 0.38, 0.37, 
  "85-0607", "a", 1.02, 1.02, 1.02, 1.01, 1.01, 
  "85-0607", "d2", 0.10, 1.21, 20.90,  NA, NA,
  "85-0607", "G", 0.53, 0.48, 0.40, 0.33, 0.25, 
  "85-0809", "a", 1.02, 1.02, 1.02, 1.01, 1.00, 
  "85-0809", "d2", 0.12, 1.27, 19.24,  NA, NA,
  "85-0809", "G", 0.44, 0.49, 0.38, 0.46, 0.28, 
  "85-1011", "a", 1.04, 1.02, 1.03, 1.01, 1.00, 
  "85-1011", "d2", 0.10, 1.48, 21.72,  NA, NA,
  "85-1011", "G", 0.51, 0.53, 0.43, 0.47, 0.32, 
  "85-C", "a", 1.02, 1.02, 1.01, 1.02, 1.02, 
  "85-C", "d2", 0.12, 1.28, 17.74,  NA, NA,
  "85-C", "G", 0.47, 0.41, 0.31, 0.30, 0.30, 
  "9192-0405", "a", 1.02, 1.02, 1.01, 1.01, 1.00, 
  "9192-0405", "d2", 0.12, 1.45, 20.68,  NA, NA,
  "9192-0405", "G", 0.42, 0.51, 0.41, 0.36, 0.29, 
  "9192-0607", "a", 1.03, 1.05, 1.02, 1.01, 1.01, 
  "9192-0607", "d2", 0.12, 1.52, 19.42,  NA, NA,
  "9192-0607", "G", 0.46, 0.43, 0.43, 0.37, 0.27, 
  "9192-0809", "a", 1.03, 1.04, 1.01, 1.01, 1.00, 
  "9192-0809", "d2", 0.11, 1.44, 17.69,  NA, NA,
  "9192-0809", "G", 0.47, 0.51, 0.44, 0.46, 0.41, 
  "9192-1011", "a", 1.02, 1.03, 1.01, 1.00, 1.00, 
  "9192-1011", "d2", 0.13, 1.64, 15.32,  NA, NA,
  "9192-1011", "G", 0.49, 0.52, 0.47, 0.43, 0.32, 
  "9192-C", "a", 1.04, 1.02, 1.03, 1.01, 1.01, 
  "9192-C", "d2", 0.12, 1.41, 14.50,  NA, NA,
  "9192-C", "G", 0.46, 0.44, 0.43, 0.39, 0.28, 
  "97-0809", "a", 1.06, 1.04, 1.01, 1.01, 1.00, 
  "97-0809", "d2", 0.12, 1.81, 20.35,  NA, NA,
  "97-0809", "G", 0.46, 0.52, 0.46, 0.47, 0.45, 
  "97-1011", "a", 1.11, 1.29, 1.01, 1.01, 1.04, 
  "97-1011", "d2", 0.11, 1.46, 18.49,  NA, NA,
  "97-1011", "G", 0.48, 0.56, 0.46, 0.42, 0.41, 
  "97-C", "a", 1.02, 1.01, 1.03, 1.08, 1.00, 
  "97-C", "d2", 0.12, 1.40, 18.63,  NA, NA,
  "97-C", "G", 0.39, 0.36, 0.44, 0.37, 0.37, 
)

summarize_glebocki <- function(data) {
  data |>
    dplyr::group_by(parameter) |>
    dplyr::summarize(across(where(is.numeric), ~round(mean(.x), 2))) |>
    dplyr::mutate(source = "glebocki", .before = 1)
}

params2 <- dplyr::bind_rows(params, summarize_glebocki(glebocki)) |> 
  dplyr::arrange(parameter)
```

A study of Sierra Nevada conifers found these values to vary by species, and
estimates of fuel loading ranged from 40% less to 8% more than when using those
suggested by @brownHandbookInventoryingDowned1974. A study of young, treated
redwood and Douglas-fir stands [@glebockiFuelLoadingMoisture2015] found similar
values for G to those of @brownHandbookInventoryingDowned1974, and values for a
much closer to those of @vanwagtendonkPhysicalPropertiesWoody1996
(@tbl-params).

```{r}
#| echo: false
#| label: tbl-params
#| tbl-cap: >
#|   Physical properties of woody fuel particles used for calculating fuel
#|   loading from several different sources in the literature.

params2 |> knitr::kable()
```


