---
title: Data description
---

# Fuel data

The fuel data was gathered in the spring of 2022 and 2023. The Firemon protocol
was used as the basis for data collection which in turn employs the use of
Brown's transects, with the addition of vegetation sample plots in which
vegetative fuel loading is estimated within a 1 meter radius cylinder.

The notion of fuel bed depth is notably absent from the [Firemon protocol], so we
added it by estimating an "average" fuel bed depth within the vegetation
sampling cylinder that included litter, and all downed woody debris.

Dominant (well represented) woody speices were also recorded in the vegetation
cylinders.

## Data entry

Data was entered using Excel and exported as a UTF-8 csv. The file format
(@fig-excel) was defined to resemble the physical datasheets as closely as
possible, while maintaining data integretity and facilitating ease of input.
There is one file for each macro plot (or replicate = site + treatment
combination). 

Each file includes sections for different tables, and these are separated by a
line that contains only a single hashtag (e.g., "#duff_litter_fbd") in lowercase
letters with underscores for spaces. In Excel, the #hashtag should go in the
first cell of a row, with nothing esle in the row. There should be no blank rows
before or after hashtags. The table sections used don't correlate 1:1 with
physical datasheet tables. For instance, columns for "metermeark1" and
"metermark2" were added to reflect the fact initially, we sampled vegetation at
5 and 10 m, but later we moved this sampling to 5 and 9 m. The "1" and "2"
following repeated variable names refers to one of these locations along the
transect. A complete description of data varialbes can be found
[below](#data-description)

- `#site_info`
  - site
  - treatment
  - date
  - onehr
  - tenhr
  - hundhr
  - thoushr
  - trans_count
  - notes *site level notes, should not contain commas*

- `#transects`
  - transect *number 1 through 8, corresponds with transect column in other
    sections*
  - corner
  - azi
  - fwd_crew
  - veg_crew *moved from vegetation table to here*
  - slope
  - metermark1 *location along transect of first veg station*
  - metermark2 *location along transect of second veg station*
  - notes *transect specific notes*

- `#duff_litter_fbd`
  - transect *number 1 through 8*
  - onehr
  - tenhr
  - hundhr
  - duff_litter1
  - pct_litter1
  - fbd1
  - duff_litter2
  - pct_litter2
  - fbd2

- `#vegetation`
  - transect *number 1 though 8*
  - live_woody1
  - dead_woody1
  - avg_w_ht1
  - live_herb1
  - dead_herb1
  - avg_h_ht1
  - live_woody2
  - dead_woody2
  - avg_w_ht2
  - live_herb2
  - dead_herb2
  - avg_h_ht2

- `#woody_species`
  - transect
  - species1 *space-delimited list of 'significant' species*
  - species2 *space-delimited list of 'significant' species*

- `#coarse_woody_debris`
  - transect
  - dia
  - decay

![Screenshot of an examle of the datasheet entry format. Each section, defined
by a begining hashtag is parsed by an r function as a seperate
table.](./figures/excel_data_example.jpg){#fig-excel}

An R function is used to parse this file into individual tables. As part of this
process, all the table sections whose rows have a 1:1 relationship with
transects are combined into one wide table. 

```{r}
#| file: "./scripts/process_datasheets.r"
```

The resulting data looks like this

```{r}
wrangle_datasheet("../data/fuel_waldon_gs.csv")
```

While the wide format for the transect data is not ideal, it is convenient because
we not working with many linked tables. This requires extra work at analysis time
to expand data stored in column names. By this, I am refering to the columns
which are followed by a number (1 or 2) which describe the location along
the transect where the data was taken. To analze these data, they will first need to
be pivoted into a longer format, where the station (1 or 2) becomes an explicit
column.

## Data variable descriptions

The following is a descripton of the data variables for each of the tables in
the data list. So, the `$plots` heading refers to the table found in
`data$plots`. This data structure is subject to change, but for now, has 3
tables. [`$plots`](#plots) has plot level data including the lengths of
transects (which are the same across all transects, but included for clarity).
[`$transects`](#transects) contains all the fuel data associated with a given
transect, including both vegetation stations, in wide format (one row for each
transect). Finally, [`$coarse_woody`](#coarse_woody) contains coarse woody
debris in a long format (multiple rows for each transect).

### The vegetation sampling cylinder

Reference is made the vegetation sampling cylinder below. It is an imaginary,
vertical cylinder with a radius of 1 meter and a variable height equal to the
maxiumm height of sprout or shrub vegetation within the cylinders radius. There
are two vegetation sampling cylinders on each transect and their centers are
defined by the transects two metermarks. Duff, litter, and fuel bed depth in
addition to the vegetation measurements, are all recorded within these
cylinders.

### `$plots`

|Variable    |Description | 
|:-----------|:-----------| 
|site        |            | 
|treatment   |            | 
|date        |            | 
|onehr       |Distance from end of transect for which 1-hr fuels were counted| 
|tenhr       |Distance from end of transect for which 10-hr fuels were counted| 
|hundhr      |Distance from end of transect for which 100-hr fuels were counted| 
|thoushr     |Distance from end of transect for which 1000-hr fuels were counted| 
|trans_count |number of transects on macro plot| 
|notes       |            | 

### `$transects`

|Variable     |Description |
|:------------|:-----------|
|site         |One of four different sites: whiskey, waldon, waldos, and camp6|
|treatment    |One of HD=High density, dispersed retention, LD=Low density dispersed retention, MD=Medium densty dispersed retention, HA=High density aggreagated retention, GS=Group selection opening (1 ha, centered on plot)|
|corner       |One of n,s,e,w for "diamond" plost and one of ne, nw, se, sw for "square" plots|
|azi          |actual azimuth from corner to end of fuel transect, deg|
|fwd_crew     |Initials of person performing fuel counts and measuring litter, duff, and FBD|
|veg_crew     |Initials of person estimating vegetation cover|
|slope        |slope in percent, measured with clinometer|
|metermark1   |locantion along transect of first veg. station m|
|metermark2   |locantion along transect of first veg. station m|
|notes        |Transect specific notes|
|onehr        |Count of down woody fuels <0.6 cm for the lenght of the 1-hr transect, redwood leaflets less than about 2 mm were not counted as one hour fuels|
|tenhr        |Count of down woody fuels >= 0.6 and < 2.5 cm|
|hundhr       |Count of down woody fuels >= 2.5 and < 8 cm|
|duff_litter1 |Combined duff and litter depth from a representative location within a 1-meter radius circle centered at metermark, cm|
|pct_litter1  |Percent of duff_litter depth compose of litter: percent|
|fbd1         |Estimated average height of litter and downed woody debris within the vegetation sampling cylinder, 1 m radius, centered at metermark|
|duff_litter2 ||
|pct_litter2  |            |
|fbd2         |            |
|live_woody1  |Total projected ground cover of all live woody plant parts within 1-meter-radius sampling cylinder of height equal to the height of the shrub or sprout vegetaion within the cylinders radius, 0-100 percent|
|dead_woody1  |Total projected ground cover of all dead woody plant parts connected to live or standing dead plants, within the vegetation sampling cylinder 0-100 percent|
|avg_w_ht1    |Average maximum height of all live and dead woody plants in vegetation sampling cylinder(see Estimating Height in Firemon protocol)|
|live_herb1   |Total projected ground cover of live herbs in the vegetation sampling cylinder|
|dead_herb1   |Total projected ground cover of dead herbs in the vegetation sampling cylinder|
|avg_h_ht1    |Average height of live and dead herbs in the vegetation sampling cylinder|
|live_woody2  |            |
|dead_woody2  |            |
|avg_w_ht2    |            |
|live_herb2   |            |
|dead_herb2   |            |
|avg_h_ht2    |            |
|species1     |Dominant woody species within sampling cylinder, each species is assumed to occupy an equal portion of volume, under-represented species are ignored|
|species2     |            |

### `$coarse_woody`

|Variable  |Description |
|:---------|:-----------|
|site      |            |
|treatment |            |
|corner    |            |
|azi       |            |
|dia       |            |
|decay     |Decay class 1-5, 1=Fine branches still present, 2=Some branches and bark missing, 3=most branches and much bark missing potentially minor decay, 4=Significant decay, 5=Almost completely rotten|


[Firemon protocol]: https://www.frames.gov/firemon/sampling-methods
