---
title: Data description
---

## Fuel data

The fuel data was gathered in the spring of 2022 and 2023. The Firemon protocol
was used as the basis for data collection which in turn employs the use of
Brown's transects, with the addition of duff, litter, and vegetation density
sampling stations at two locations along each transect
(@sec-sampling-cylinder). While the firemon protocol specifies a 2 meter tall
imaginary cylinder for quantifying surface fuels, because our resprouting trees
were continuous with the ground, we used an imaginary cylinder with a variable
height, up to the tallest shrub or sprout fuels that were continuous to the
ground.

Two 10-meter transects were installed parallel to plot edges at each macro plot
corner. Their near ends shared a point inset 10 meters from the corner, towards
plot center (@fig-sampling-diagram). Downed, woody fuels in four size classes
were counted starting from the far end of the transect and each fuel size class
was counted for a specified lenght of the total transect (generally longer for
larger fuels).

At two stations along each transect, duff, litter, live and dead vegetation
density (as height and percent cover), and fuel bed depth were assessed. Details
of each of these values are described below (@sec-sampling-cylinder)

The notion of fuel bed depth is notably absent from the [Firemon protocol], so
we added it by estimating an "average" fuel bed depth within the vegetation
sampling cylinder that included litter, and all downed woody debris.

Dominant (well represented) woody speices were also recorded in the vegetation
cylinders.

![A diagram of the fuel sampling design depicting 0.4 hectare macro plot, fuels
transects, transect stations, and vegetation sub-plots. Large green circles are
vegetation monitoring plots (which are mostly used for tracking species
composition) and the small orange circles are transect stations where duff,
litter, and live and dead vegetation density were
measured.](./figures/fuel_sampling_layout.png){#fig-sampling-diagram}

```{r}
#| label: fig-sampling-design
#| fig-cap: >
#|   Diagrammatic representation of the experimental design. Black points
#|   are fuel transect corners, and line segments represent individual
#|   transects. The outline represent the macro-plot boundaries.
#| code-fold: true
#| message: false

library(ggplot2)
library(gt)
library(dplyr)

draw_sampling_design <- function() {
  max <- 44.7
  # CW from bottom left
  xdir <- c(1, 1, -1, -1)
  ydir <- c(1, -1, -1, 1)
  corner_inset <- 10 / sqrt(2)
  centerx <- xdir * corner_inset + c(0, 0, max, max)
  centery <- ydir * corner_inset + c(0, max, max, 0)
  xend <- xdir * 10 + centerx
  yend <- ydir * 10 + centery
  centers <- data.frame(centerx, centery)
  ends <- rbind(centers, centers)
  # Half of the x and y coordinates of the transect ends are the same as the
  # center, and half are different.
  ends$xend <- c(xend, centerx)
  ends$yend <- c(centery, yend)

  data <- tidyr::expand_grid(
    treatment = c("GS", "LD", "HA", "HD"),
    site = factor(c(1, 2, 3, 4)), 
    ends
  )
  
  p <- ggplot(data) +
  geom_rect(
    aes(xmin = 0, ymin = 0, xmax = 44.7, ymax = 44.7),
    color = "black",
    fill = NA
  )+
  geom_point(aes(centerx, centery)) +
  geom_segment(aes(centerx, centery, xend = xend, yend = yend)) +
  facet_grid(
    site ~ treatment,
    switch = "y"
  ) +
  theme(
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.spacing = unit(0, "in"),
    aspect.ratio = 1
  ) +
  scale_x_continuous(position = "top") +
  labs(x = "Treatment", y = "Site")
  print(p)
}

draw_sampling_design()

```

### Data entry {#sec-data-entry}

Data was entered using Excel and exported as a UTF-8 csv. The file format
(@fig-excel) was defined to resemble the physical datasheets as closely as
possible, while maintaining data integretity and facilitating ease of input.
There is one file for each macro plot (or replicate = site + treatment
combination). 

Each file includes sections for different tables, and these are separated by a
line that contains only a single hashtag (e.g., "#duff_litter_fbd") in lowercase
letters with underscores for spaces. In Excel, the #hashtag should go in the
first cell of a row, with nothing esle in the row. There should be no blank rows
before or after hashtags. The table sections used don't correlate 1:1 with
physical datasheet tables. For instance, columns for "metermeark1" and
"metermark2" were added to reflect the fact that initially we sampled duff,
litter and vegetation density (transect stations) at 5 and 10 m, but later we
moved the stations to 5 and 9 m. The "1" and "2" following repeated variable
names refers to one of these locations along the transect. The function used to
wrangle these data matches rows based on transect number and columns based on
column names. Thus, it is possible to rearrage the rows or columns within each
section as long as column names are kept consistent and data for each transect
have the same transect number. Following are the section and column labels used
for each section of the plot data entry form with any notes regarding their
format. A complete description of the data varialbes can be found in
@sec-sampling-cylinder. 

- `#site_info` site treatment
  - date
  - onehr
  - tenhr
  - hundhr
  - thoushr
  - trans_count notes *(site level notes, should not contain commas)*

- `#transects`
  - transect *(number 1 through 8, corresponds with transect column in other
    sections)*
  - corner
  - azi
  - fwd_crew
  - veg_crew *(moved from vegetation table to here)*
  - slope
  - metermark1 *(location along transect of first station)*
  - metermark2 *(location along transect of second station)*
  - notes *(transect specific notes)*

- `#duff_litter_fbd`
  - transect *(number 1 through 8)*
  - onehr
  - tenhr
  - hundhr
  - duff_litter1
  - pct_litter1
  - fbd1
  - duff_litter2
  - pct_litter2
  - fbd2

- `#vegetation`
  - transect *(number 1 though 8)*
  - live_woody1
  - dead_woody1
  - avg_w_ht1
  - live_herb1
  - dead_herb1
  - avg_h_ht1
  - live_woody2
  - dead_woody2
  - avg_w_ht2
  - live_herb2
  - dead_herb2
  - avg_h_ht2

- `#woody_species`
  - transect
  - species1 *(space-delimited list of 'significant' species)*
  - species2 *(space-delimited list of 'significant' species)*

- `#coarse_woody_debris`
  - transect
  - dia
  - decay

![Screenshot of an examle of the datasheet entry format. Each section, defined
by a begining hashtag is parsed by an r function as a seperate
table.](./figures/excel_data_example.jpg){#fig-excel}

An R function is used to parse this file into individual tables. As part of this
process, all the table sections whose rows have a 1:1 relationship with
transects are combined into one wide table, based on the transect column.

```{r}
#| code-fold: true
#| file: "./scripts/process_datasheets.r"
```

The resulting data looks like this:

```{r}
wrangle_datasheet("../data/fuel_waldon_gs.csv")
```

While the wide format for the transect data is not ideal, it is convenient
because it reduces the number of tables we are dealing with. This will require
extra work at analysis time to expand the data variables which are followed by a
"1" or "2", which represent different stations (@sec-sampling-cylinder) on the
same transect. To analze these data, they will first need to be pivoted into a
longer format, where the station (1 or 2) becomes an explicit column.

### Data variable descriptions

The following is a descripton of the data variables for each of the tables in
the data list. So, the `$plots` heading refers to the table found in
`data$plots`. This data structure is subject to change, but for now, has 3
tables. [`$plots`](#plots) has plot level data including the lengths of
transects (which are the same across all transects, but included for clarity).
[`$transects`](#transects) contains all the fuel data associated with a given
transect, including both sampling stations, in wide format (one row for each
transect). Finally, [`$coarse_woody`](#coarse_woody) contains coarse woody
debris in a long format (multiple rows for each transect).

### Station sampling cylinder {#sec-sampling-cylinder}

Reference is made the station sampling cylinder below. It is an imaginary,
vertical cylinder with a radius of 1 meter and a variable height equal to the
maxiumm height of sprout or shrub vegetation within the cylinders radius. There
are two sampling cylinders on each transect and their centers are defined by the
transects' two metermarks. Duff, litter, and fuel bed depth in addition to the
vegetation measurements, are all recorded within these cylinders.

### `$plots`

|Variable    |Description | 
|:-----------|:-----------| 
|site        |            | 
|treatment   |            | 
|date        |            | 
|onehr       |Distance from end of transect for which 1-hr fuels were counted| 
|tenhr       |Distance from end of transect for which 10-hr fuels were counted| 
|hundhr      |Distance from end of transect for which 100-hr fuels were counted| 
|thoushr     |Distance from end of transect for which 1000-hr fuels were counted| 
|trans_count |number of transects on macro plot| 
|notes       |            | 

### `$transects`

|Variable     |Description |
|:------------|:-----------|
|site         |One of four different sites: whiskey, waldon, waldos, and camp6|
|treatment    |One of HD=High density, dispersed retention, LD=Low density dispersed retention, MD=Medium densty dispersed retention, HA=High density aggreagated retention, GS=Group selection opening (1 ha, centered on plot)|
|corner       |One of n,s,e,w for "diamond" plost and one of ne, nw, se, sw for "square" plots|
|azi          |actual azimuth from corner to end of fuel transect, deg|
|fwd_crew     |Initials of person performing fuel counts and measuring litter, duff, and FBD|
|veg_crew     |Initials of person estimating vegetation cover|
|slope        |slope in percent, measured with clinometer|
|metermark1   |locantion along transect of first veg. station m|
|metermark2   |locantion along transect of first veg. station m|
|notes        |Transect specific notes|
|onehr        |Count of down woody fuels <0.6 cm for the lenght of the 1-hr transect, redwood leaflets less than about 2 mm were not counted as one hour fuels|
|tenhr        |Count of down woody fuels >= 0.6 and < 2.5 cm|
|hundhr       |Count of down woody fuels >= 2.5 and < 8 cm|
|duff_litter1 |Combined duff and litter depth from a representative location within a 1-meter radius circle centered at metermark, cm|
|pct_litter1  |Percent of duff_litter depth compose of litter: percent|
|fbd1         |Estimated average height of litter and downed woody debris within the sampling cylinder, 1 m radius, centered at metermark|
|duff_litter2 ||
|pct_litter2  |            |
|fbd2         |            |
|live_woody1  |Total projected ground cover of all live woody plant parts within 1-meter-radius sampling cylinder of height equal to the height of the shrub or sprout vegetaion within the cylinders radius, 0-100 percent|
|dead_woody1  |Total projected ground cover of all dead woody plant parts connected to live or standing dead plants, within the sampling cylinder 0-100 percent|
|avg_w_ht1    |Average maximum height of all live and dead woody plants in sampling cylinder (see Estimating Height in Firemon protocol)|
|live_herb1   |Total projected ground cover of live herbs in the sampling cylinder|
|dead_herb1   |Total projected ground cover of dead herbs in the sampling cylinder|
|avg_h_ht1    |Average maximum height of live and dead herbs in the sampling cylinder|
|live_woody2  |            |
|dead_woody2  |            |
|avg_w_ht2    |            |
|live_herb2   |            |
|dead_herb2   |            |
|avg_h_ht2    |            |
|species1     |Dominant woody species within sampling cylinder, each species is assumed to occupy an equal portion of volume, under-represented species are ignored|
|species2     |            |

### `$coarse_woody`

|Variable  |Description |
|:---------|:-----------|
|site      |            |
|treatment |            |
|corner    |            |
|azi       |            |
|dia       |            |
|decay     |Decay class 1-5, 1=Fine branches still present, 2=Some branches and bark missing, 3=most branches and much bark missing potentially minor decay, 4=Significant decay, 5=Almost completely rotten|


[Firemon protocol]: https://www.frames.gov/firemon/sampling-methods
