---
title: Fuel data exploration
---

Our main questions regarding fuel loading are:

1. Are there differences in fuel loading between treatments?
2. If so, what is the magnitude of the difference?

We have a range of response variables that include loading in several different
classes of surface fuels. Some of these may be correlated with each other, which
may have an effect on our interpretation of differences between treatments. For
instance, if duff-litter load is negatively correlated with vegetation, even if
we don't see differences between treatment in either of these variables
separately, it might be found that for any given level of vegetation loading, one
treatment, or another may have consistently higher levels of duff-litter.
Capturing differences between treatments in terms of interactions like these
requires different tools than assessing any one variable on its own.

In order to assess potential outliers, co-linearity, interactions, and other
problems with statistical tests, we'll first conduct some data exploration as
outlined by @zuurProtocolDataExploration2010.

For reference, here is a list of the fuel loading (response) variables of
interest.

1. duff/litter load
2. woody vegetation
3. herbaceous vegetation
4. one hr fuels
5. ten hr fuels
6. hundred hr fuels
7. fine woody debris (one, ten, and hundred hr fuels)
8. coarse woody debris

It's important to note that variables 1-3 were measured twice per transect and
the rest were measured once per transect.

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
load("./data_long.RData")
transectid <- c("site", "treatment", "corner", "azi")
```

## Outliers

```{r}
#| warning: false
#| message: false

cleavland_plot <- function(data, title, load_var = load) {
  data |>
    group_by(site, treatment) |>
    mutate(replicate_mean_load = mean({{load_var}}, na.rm = TRUE)) |>
    ggplot(
      aes(
        {{load_var}},
        fct_reorder( interaction(site, treatment, sep = " "), {{load_var}}),
        color = treatment)
    ) +
    geom_jitter(width = 0, height = 0.2) +
    labs(x = expression(Load~(Mg%.%ha^-1)), y = "Data order", title = title)
}

cwd |>
  summarize( load = sum(load, na.rm = TRUE), .by = all_of(transectid)) |> 
  cleavland_plot("CWD")

dufflitter |>
  cleavland_plot("Duff/litter")

fwd |> 
  group_by(class) |> 
  group_walk(\(data, groups) print(cleavland_plot(data, groups$class)))

veg |>
  group_by(class) |>
  group_walk(\(data, groups) print(cleavland_plot(data, groups$class)))

```
