---
title: Reneration data summary
---

```{r}
#| output: false

library(tidyverse)
library(ggdist)

load("regen_wrangled.RData")
```

## Objectives

- Determine if there are significant differences in species composition, or size
  class distirbution between treatments

This objective is somewhat vaugue because I am not familiar with the statistical
techniques used for the weighted (diameter), multi-variate (species)
distribution data that we have.

For starters, I will look at summaries of size class distributions by species. I
will report densities in terms of trees per hectare (tph). Our plots were fixed
area, 4-m-radius plots, or 50.265 m^2^. Each stem represents 198.946 stems/ha.
Or if we sum at the plot level (4 veg plots), then 49.736 stems/ha, or 12.434
stems/ha at the treatment level (16 veg plots).

```{r}
summary(regen)

hist(regen$dbh)

filter(regen, dbh > 2) |> pull() |> hist()

regen |>
  mutate(
    bin = cut(dbh, 10)
  ) |>
  ggplot(aes(bin)) +
  geom_histogram(aes(stat = "count"))

regen |>
  filter(str_detect(spp, "NETTLE")) |>
  ggplot(aes(dbh, fill = spp, color = spp)) +
  geom_histogram()

```


