---
title: Data import and description
format:
  html:
    code-fold: true
---

<!--
source(knitr::purl("sprouts_import_data.qmd", tempfile()))
-->

```{r}
#| include: false

# Quickly iterate over random color selections and pick the colors I like
rnd_color_brewer <- function(palette, chosen_order = NULL) {
  n <- RColorBrewer::brewer.pal.info[palette, "maxcolors"]
  ord <- sample(1:n)
  if(!is.null(chosen_order)) ord <- chosen_order
  colors <- RColorBrewer::brewer.pal(n, palette)[ord]
  print(ord)
  colors
}

ggplot2::theme_update(
  panel.grid.major = ggplot2::element_blank(),
  panel.grid.minor = ggplot2::element_blank()
)
```

## Description

First, we'll load some libraries and our data.

```{r}
#| label: import-data

suppressWarnings(suppressPackageStartupMessages(library(tidyverse)))
library(ggridges)
library(ggokabeito)

sprouts_fresh <- readxl::read_excel("../data/Sprouts10yr.xlsx")

set.seed(743)
slice_sample(sprouts_fresh, n = 4) |> knitr::kable()
```

The data is in a wide format, variables are as follows:

```{r}
#| label: tbl-variable-descriptions
#| tbl-cap: Descriptions of variables in dataset.
#| echo: false

variable_descriptions <- list(
  Site = "One of four sites where treatments were replicated. Sites were located
  on similar slope positions, but across a range of aspects.",
  Plot = "There were 4 plots at each site and each was randomly assigned a
  treatment",
  Trtmt = "Treatemnt type: GS = group selection, which is basically a small
  clearing, LD = low density--fewer trees remaining; HD = high density--more
  trees remain and they are dispersed; HA = high density aggregated--more trees
  remain and they are grouped into clumps.",
  Tree = "Unique sprout ID within Site, Plot, and Species",
  Species = "SESE = coast redwood, and LIDE = tanoak.",
  HT1yr_m = "Sprout height one year after treatment.",
  HT5yr_m = "Same as above, but for year 5",
  HT10yr_m = "Same as above, but for year 10",
  `HTI1-5yr` = "Height growth between years 1 and 5, (4 growth periods)",
  `HTI5-10yr` = "Height growth between years 5 and 10 (5 growth periods)",
  DBH10yr_cm = "Diameter at breast height in cm at year 10. Only collected for
  redwood",
  LCBH10yr = "Live crown base height (height to first live branch) at year 10. 
  Only collected for redwood.",
  CR10yr = "Crown ratio (live crown length / total height) at year 10, only for
  redwood",
  HD10yr = "Unknown.",
  SDIinit = "Stand density index of plot immediately after treatment.",
  AggregatedYN = "Indicator for treatment HA.",
  ResidRWonClumpYN = "Unknown.",
  HT10rank = "Trees species specific height ranking within plot."
)

variable_descriptions |>
  imap(function(x, y) {
    tibble(Variable = y, Description = gsub("[[:space:]]{2,}", " ", x))
  }) |>
  list_rbind(names_to = "Variable") |>
  knitr::kable()

```

## Wrangle

I'm going to change some variable names to make them more ergonomic

```{r}

newnames <- c(
  site = "Site", plot = "Plot", treat = "Trtmt", tree = "Tree", spp = "Species",
  ht1 = "HT1yr_m", ht5 = "HT5yr_m", ht10 = "HT10yr_m", ht_inc5 = "HTI1-5yr",
  ht_inc10 = "HTI5-10yr", dbh10 = "DBH10yr_cm", lcbh10 = "LCBH10yr",
  sdi_init = "SDIinit", agg = "AggregatedYN", ht_rnk10 = "HT10rank"
)

tibble(old = newnames, new = names(newnames)) |> knitr::kable()

```

Our data are nested. We have plots within sites, trees within plots, and
observeations within trees (multiple observations per tree).

```
Sites
└─ Plots
   └─ Trees
      └─ Observations
```

Each Treatment is represented by one site/plot combination. Each site belongs to
each treatment and vice versa. I think the terminology here is that Sites and
treatments are crossed.

Currently, `plot` (integer) is only unique within `site` and `tree` is only unique within
`site`, `treat`, and `spp`. It will be more convenient if `plot` and `tree` are
globally unique identifiers. This makes the nesting structure implicit and we
can simplify our model syntax.

```{r}
#| output: false

make_site_plot_unique <- function(data) {
  data |>
    mutate(plot = cur_group_id(), .by = c(site, treat)) |>
    mutate(tree = row_number())
}

```

I'm also going to order the treatments acording to our expectations about the
most to least productive. This will affect how they are plotted and reported.

```{r}

treat_order <- c("GS", "LD", "HA", "HD")

set_expected_treatment_order <- function(data) {
  data |>
    mutate(treat = fct_relevel(treat, treat_order))
}

```

And I'll apply all the above steps in one go.

```{r}

sprouts <- sprouts_fresh |>
  select(all_of(newnames)) |>
  make_site_plot_unique() |>
  set_expected_treatment_order()

```

I will need a function to convert a variable in the data from wide to long format.

```{r}

lengthen_data <- function(data, var) {
  # I need regex for first part to handle ht and ht_inc and
  pref <- switch(var, ht = "(ht)", ht_inc = "(ht_inc)")
  # and the year
  suf <- "(\\d+)"
  str_to_match <- paste0(pref, suf)
  pivot_longer(data,
    matches(str_to_match),
    names_to = c(".value", "year"),
    names_pattern = str_to_match,
    names_transform = list(year = as.integer)
  ) |>
  relocate(year, matches(paste0(pref, "$")), .after = spp)
}

```

## Objectives

Our main objectives are to:

- Test for treatment effects on redwood and tanoak height (or height increment)
  over time, but especially for year 10
- Explore treatment effect on competition between redwood and tanoak
- Produce a model for sapling height as a funtion of harvest intensity (initial 
  SDI) and years since harvest

## Visualize data

### Tree heights

The following figures reveal possible trends in the raw data.

#### Species/treatment/year

@fig-data-summary-treatments reveals fine, species specific differences between
treatments and general trends over time. It shows that over time, LD and GS
treatments have more taller trees than the other treatments, with GS having the
most. This is true for both redwood and tanoak. Additionaly, for redwood it
appears that HA may have taller trees than HD, at least in year 10.

@fig-data-summary-spp shows the same information as before, but is arguably 
easier to look at. There is a generall increasing trend in heights with
treatments as follows:

    HD < HA < LD < GS

Redwood is consistently taller than tanoak, and the GS treatment confers the
greatest advantage to redwood.

Across all treatments, it is also interesting to note that over times, the
height distributions, especially for redwood, seem to becoming more multi-modal
and more widely distributed. This could be due to site or plot effects, or to
microsite (within plot) effects, but it is not immediately clear why this
diverging performance should be so apparent with redwood and not tanoak. 


```{r}
#| label: fig-data-summary-treatments
#| fig-width: 8
#| fig-cap: >
#|   Height distributions for tanoak and redwood over time for each of four
#|   treatments. Treatments GS and LD have higher proportions of taller trees.
#|   Years refers to number of years after treatment.
#| message: false

sprouts |>
  lengthen_data("ht") |>
  ggplot(aes(ht, fct_rev(factor(year)), color = treat)) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  geom_density_ridges(alpha = .3, size = 1) +
  facet_wrap(~spp) +
  labs(x = "Height (m)", y = "Year") +
  scale_color_brewer(palette = "Set2")
```

```{r}
#| label: fig-data-summary-spp
#| fig-height: 8
#| fig-cap: >
#|   Similar to @fig-data-summary-treatments, but with an emphasis on
#|   differences between speceis responses across treatments. Comparisons are
#|   made for each year. The GS treatment appears to favor redwood response the
#|   most, but all treatments show redwoods are taller than tanoak. Years refers
#|   to number of years after treatment.
#| message: false

sprouts |>
  lengthen_data("ht") |>
  ggplot(aes(
    x = ht,
    y = fct_relevel(treat, treat_order),
    color = factor(spp, levels = c("SESE", "LIDE"))
  )) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  geom_density_ridges(alpha = .3, size = 1) +
  facet_grid(year ~ ., labeller = label_both) +
  labs(x = "Height (m)", y = "Treatment") +
  scale_color_brewer(palette = "Set2", name = "Species")
```

#### Species/year/SDI
@fig-data-summary-sdi shows that above around 400 SDI, tree heights level off.
It also implies a steady decrease in height from 0 to around 400 SDI. The
strength of the relationship appears to be increasing over time, particularly
for redwood.

```{r}
#| label: fig-data-summary-sdi
#| fig-cap: SDI vs HT for both species, across measurement years.
#| fig-width: 8.5

sprouts |>
  lengthen_data("ht") |>
  ggplot(aes(sdi_init, ht, color = factor(year))) +
  geom_point(alpha = 0.5) +
  geom_smooth(span = 1.2) +
  facet_wrap(~spp) +
  scale_color_brewer(palette = "Set2")
```

#### Species/site/plots
Lets see what the variability among sites and plots looks like, I'll focus on
year 10 only.

@fig-data-summary-sites reveals some differences between sites, particularly for
redwoods. Waldo North tends to have larger redwoods and Camp 6 has a large
proportion of smaller redwoods.

We can see in @fig-data-summary-plots that there seems to be differences among
plots beyond site differences and treatment differences. We should expect plots
to capture a portion of the variance. Most notable is the plot level difference
between redwood and tanoak. For redwood, the large ammount of within plot
variability combined with the between plot variability obscures the treatment
effect. If you squint, there appears to be a similar overall pattern between
redwood and tanoak repsonse to treatment, but it appears they respond
differentially to certain plots.

```{r}
#| label: fig-data-summary-sites
#| fig-cap: >
#|   Distribution of heights at year 10 at each site for two species. Slabs are
#|   normalized by sample size to reflect the raw data, plotted as dots below.

suppressPackageStartupMessages(library(ggdist))

sprouts |>
  ggplot(aes(ht10, site, fill = spp)) +
  stat_slab(aes(thickness = after_stat(pdf * n)), alpha = 0.6, scale = 0.7) +
  stat_dotsinterval(side = "bottom", scale = 0.7, slab_color = NA) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "year 10 height (m)", y = NULL, fill = "species")

```

```{r}
#| label: fig-data-summary-plots
#| fig-height: 7
#| fig-cap: >
#|   Same as for @fig-data-summary-sites, but for each plot (Site/treatment
#|   interaction). Grouped by treatment.

sprouts |>
  ggplot(aes(
    ht10,
    interaction(site, fct_relevel(treat, treat_order), sep = " - "),
    fill = fct_relevel(treat, treat_order),
  )) +
  stat_slab(aes(thickness = after_stat(pdf * n)), alpha = 0.5, scale = 0.7) +
  stat_dotsinterval(side = "bottom", scale = 0.7, slab_color = NA) +
  scale_color_brewer(palette = "Set2") +
  facet_wrap(~spp) +
  theme(legend.position = "bottom") +
  labs(y = NULL, x = "year 10 height (m)", fill = "treatment")

```

### Height increments

Height increments contain similar information as heights, but allow us to
compare directly between years.

@fig-data-summary-ht-inc-year shows that across treatments and species, height
growth slows down in the second period (years 5-10). This is more true for
redwood but it starts with more rapid growth than tanoak. In the most crowded
treatment (HD), redwoods height increment has become slower than tanoaks in the
second period. Also, in the second period, the high density, aggregated
treatment appears to have slightly higher (or equal) average growth increment,
which is not completely expected.

```{r}
#| label: fig-data-summary-ht-inc-year
#| fig-height: 7
#| fig-cap: >
#|   Variations in annual height growth increment between the first and second
#|   measurement periods (years 1-5, and 5-10, respectively).

sprouts |>
  lengthen_data("ht_inc") |>
  ggplot(
    aes(
      ht_inc,
      fct_relevel(treat, treat_order),
      fill = factor(year),
      color = factor(year)
    )
  ) +
  stat_slab(alpha = 0.5) +
  stat_pointinterval(
    position = position_dodge(width = 0.4, preserve = "single")
  ) +
  facet_wrap(~spp) +
  theme(legend.position = "bottom") +
  labs(
    fill = "Year",
    color = "Year",
    y = "Treatment",
    x = expression(Height~increment~(m~yr^-1))
  ) +
  scale_color_manual(
    values = rnd_color_brewer("Set2", c(1,4)),
    aesthetics = c("color", "fill")
  )

```

## Modeling
In addition to characterizing redwood growth response, we would also like to
accurately describe differences between redwood and tanoak. This is probably
best accomplished by including species in the model, and modeling the difference
directly.

because of our multiple grouping variables, model complexity can grow quickly,
especially when considering how covariates interact with grouping variables,
which lead to random slope models

I'm also wondering what it means to include a categorical variable as a random
slope (on the LHS of the random parts). Michael Clark [explores this scenario],
with some simpler data.

[explores this scenario]: https://m-clark.github.io/posts/2020-03-01-random-categorical

In any case, following the pattern of data summaries above, I'll start by
modeling to detect treatment differences in tree height.

I'll start with the simplest model that I'm willing to look at: a
treatment/species interaction for only year 10. This model is not that bad, the
residuals indicate some heteroskedasticity. It has a moderate r-squared (0.55).
It doesn't account for non-independence due to nesting structure of our data
though.

```{r}
#| code-fold: false
#| fig-cap: >
#|   Model predictions for year 10 height, with 95% confidence intervals.

# I'm calling the height data "d"
d <- sprouts |>
  lengthen_data("ht")

m1 <- lm(ht ~ treat * spp, data = filter(d, year == 10))

library(marginaleffects)

d1 <- predictions(m1, by = c("treat", "spp"))

filter(d, year == 10) |>
  ggplot(aes(treat, ht, color = spp, fill = spp)) +
  facet_grid(~spp, switch = "x") +
  theme(panel.spacing = unit(0, "lines"),
        strip.background = element_blank(),
        strip.text = element_blank(),
        ) +
  geom_dots() +
  scale_color_brewer(palette = "Set2", aesthetics = c("color", "fill")) +
  geom_pointrange(
    data = d1, 
    aes(y = estimate, ymin = conf.low, ymax = conf.high),
    color = "gray50", position = position_nudge(x = -0.07),
    size = 0.7,
    linewidth = 1
  )

```

Do we expect height to increase lineraly with year? First, lets look at individual tree's height trajectories.

```{r}

d |> 
  ggplot(aes(year, ht, color = spp, group = tree)) +
  geom_line(alpha = 0.6) +
  facet_wrap(~treat*site) +
  scale_color_brewer(palette = "Set2")

```

For the most part, it looks like height increases linearly with year. So lets
fit some models with year as a continuous variable.

```{r}
library(modelsummary)
library(lme4)

m2 <- lm(ht ~ treat * spp * year, data = d)

m3 <- lm(ht ~ treat * spp + year + treat:year + spp:year, data = d)

```


```{r}
#| code-fold: false

m4 <- lmer(
  ht ~ treat * spp * year + (1 | site) + (1 | plot) + (1 | tree),
  data = d
)
options(modelsummary_model_labels = "model")
modelsummary(list(m1, m2, m3, m4), stars = TRUE)


```


