---
title: Data import and description
---

<!--
source(knitr::purl("sprouts_import_data.qmd", tempfile()))
-->

```{r}
#| include: false

ggplot2::theme_update(
  panel.grid.major = ggplot2::element_blank(),
  panel.grid.minor = ggplot2::element_blank()
)
```

First, we'll load some libraries and our data.

```{r}
#| label: import-data

suppressWarnings(suppressPackageStartupMessages(library(tidyverse)))
library(ggridges)
library(ggokabeito)

sprouts <- readxl::read_excel("../data/Sprouts10yr.xlsx")

set.seed(743)
slice_sample(sprouts, n = 4) |> knitr::kable()
```

The data is in a wide format, variables are as follows:

```{r}
#| label: tbl-variable-descriptions
#| tbl-cap: Descriptions of variables in dataset.
#| echo: false

variable_descriptions <- list(
  Site = "One of four sites where treatments were replicated. Sites were located
  on similar slope positions, but across a range of aspects.",
  Plot = "There were 4 plots at each site and each was randomly assigned a
  treatment",
  Trtmt = "Treatemnt type: GS = group selection, which is basically a small
  clearing, LD = low density--fewer trees remaining; HD = high density--more
  trees remain and they are dispersed; HA = high density aggregated--more trees
  remain and they are grouped into clumps.",
  Tree = "Unique sprout ID within Site, Plot, and Species",
  Species = "SESE = coast redwood, and LIDE = tanoak.",
  HT1yr_m = "Sprout height one year after treatment.",
  HT5yr_m = "Same as above, but for year 5",
  HT10yr_m = "Same as above, but for year 10",
  `HT1-5yr` = "Height growth between years 1 and 5",
  `HT5-10yr` = "Height growth between years 5 and 10",
  DBH10yr_cm = "Diameter at breast height in cm at year 10. Only collected for
  redwood",
  LCBH10yr = "Live crown base height (height to first live branch) at year 10. 
  Only collected for redwood.",
  CR10yr = "Crown ratio (live crown length / total height) at year 10, only for
  redwood",
  HD10yr = "Unknown.",
  SDIinit = "Stand density index of plot immediately after treatment.",
  AggregatedYN = "Indicator for treatment HA.",
  ResidRWonClumpYN = "Unknown.",
  HT10rank = "Trees species specific height ranking within plot."
)

variable_descriptions |> 
  imap(\(x, y) tibble(Variable = y, Description = gsub("[[:space:]]{2,}", " ", x))) |>
  list_rbind(names_to = "Variable") |>
  knitr::kable()

```

Our main objectives are to:

- Test for treatment effects on redwood height and diameter at year 10
- Explore treatment effect on competition between redwood and tanoak
- Model the relationship between both of these and harvest intensity (SDIinit)
- Explore differences in height response over multiple remeasurements

The following figures reveal possible trends in the raw data.

@fig-data-summary-treatments shows that over time, LD and GS treatments have
more taller trees than the other treatments, with GS having the most. This is
true for both redwood and tanoak.

@fig-data-summary-spp seems to suggest that redwood are taller across all
treatments and years, but the GS treatment is the only one that provides a clear
advantage to redwood.

@fig-data-summary-sdi shows that above around 400 SDI, tree heights level off.
It also implies a steady decrease in height from 0 to around 400 SDI. The
strength of the relationship appears to be increasing over time, particularly
for redwood.

```{r}
#| label: fig-data-summary-treatments
#| fig-width: 8
#| fig-cap: >
#|   Height distributions for tanoak and redwood over time for each of four
#|   treatments. Treatments GS and LD have higher proportions of taller trees.
#|   Years refers to number of years after treatment.
#| message: false

# Function to convert heights from multiple years to long format
lengthen_heights <- function(data) {
  pivot_longer(data,
    matches("HT..?yr_m"),
    names_to = "year", values_to = "height",
    names_transform = \(x) as.integer(str_extract(x, "\\d+"))
  ) |>
  relocate(c(year, height), .after = Species)
}

sprouts |>
  lengthen_heights() |>
  ggplot(aes(height, fct_rev(factor(year)), color = Trtmt)) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  geom_density_ridges(alpha = .3, size = 1) +
  facet_wrap(~Species) +
  labs(x = "Height (m)", y = "Year") +
  scale_color_brewer(palette = "Set2")

```

```{r}
#| label: fig-data-summary-spp
#| fig-height: 8
#| fig-cap: >
#|   Similar to @fig-data-summary-treatments, but with an emphasis on
#|   differences between speceis responses across treatments. Comparisons are
#|   made for each year. The GS treatment appears to favor redwood response the
#|   most, but all treatments show redwoods are taller than tanoak. Years refers
#|   to number of years after treatment.
#| message: false

sprouts |>
  lengthen_heights() |>
  ggplot(aes(
    x = height,
    y = factor(Trtmt,levels = c("GS", "LD", "HA", "HD")),
    color = factor(Species, levels = c("SESE", "LIDE"))
  )) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  geom_density_ridges(alpha = .3, size = 1) +
  facet_grid(year ~ ., labeller = label_both) +
  labs(x = "Height (m)", y = "Treatment") +
  scale_color_brewer(palette = "Set2", name = "Species")

```

```{r}
#| label: fig-data-summary-sdi
#| fig-cap: SDI vs HT for both species, across measurement years.
#| fig-width: 8.5

sprouts |>
  lengthen_heights() |>
  ggplot(aes(SDIinit, height, color = factor(year))) +
  geom_point(alpha = 0.5) +
  geom_smooth(span = 1.2) +
  facet_wrap(~Species) +
  scale_color_brewer(palette = "Set2")

```

Lets see what the variability among sites and plots looks like, I'll focus on
year 10 only.

@fig-data-summary-sites reveals some differences between sites, particularly for
redwoods. Waldo North tends to have larger redwoods and Camp 6 has a large
proportion of smaller redwoods.

We can see in @fig-data-summary-plots that there seems to be differences among
plots beyond site differences and treatment differences. We should expect plots
to capture a portion of the variance.

```{r}
#| label: fig-data-summary-sites
#| fig-cap: >
#|   Distribution of heights at year 10 at each site for two species. Slabs are
#|   normalized by sample size to reflect the raw data, plotted as dots below.

library(ggdist)

sprouts |>
  ggplot(aes(HT10yr_m, Site, fill = Species)) +
  stat_slab(aes(thickness = after_stat(pdf*n)), alpha = 0.6, scale = 0.7) +
  stat_dotsinterval(side = "bottom", scale = 0.7, slab_color = NA) +
  scale_fill_brewer(palette = "Set2")

```

```{r}
#| label: fig-data-summary-plots
#| fig-cap: >
#|   Same as for @fig-data-summary-sites, but for each plot (Site/treatment
#|   interaction). Grouped by treatment.

sprouts |>
  ggplot(aes(
      HT10yr_m,
      interaction(Site, Trtmt, sep = " - "),
      fill = Trtmt
  )) +
  stat_slab(aes(thickness = after_stat(pdf * n)), alpha = 0.5, scale = 0.7) +
  stat_dotsinterval(side = "bottom", scale = 0.7, slab_color = NA) +
  scale_color_brewer(palette = "Set2")

```

Our data though are nested, we have:

```
Sites
└─ Plots
   └─ Trees
      └─ Observations
```

Each Treatment is represented by one site/plot combination. Each site belongs to
each treatment and vice versa. I think the terminology here is that Sites and
treatments are crossed.

## Data wrangling

Currently, `Plot` is only unique within `Site` and `Tree` is only unique within
`Site`, `Trtmt`, and `Species`. It will be more convenient if `Plot` and `Tree` are
globally unique identifiers. This makes the nesting structure implicit, and we
can simplify our model syntax.

```{r}
library(lme4)

implicit_grouping <- function(data) {
  data |>
    mutate(Plot = cur_group_id(), .by = c(Site, Trtmt)) |>
    mutate(Tree = row_number())
}

d <- sprouts |>
  implicit_grouping() |>
  lengthen_heights()
```

## Model 1

In addition to characterizing redwood growth response, we would also like to
accurately describe differences between redwood and tanoak. This is probably
best accomplished by including species in the model, and modeling the difference
directly.

because of our multiple grouping variables, model complexity can grow quickly,
especially when considering how covariates interact with grouping variables,
which lead to random slope models

I'm also wondering what it means to include a categorical variable as a random
slope (on the LHS of the random parts). Michael Clark [explores this scenario],
with some simpler data.

[explores this scenario]: https://m-clark.github.io/posts/2020-03-01-random-categorical

```{r}

m <- lmer(
  height ~ scale(SDIinit) * factor(year) * Species + (1 | Site) + (1 | Plot) + (1 | Tree),
  # data = subset(d, Species == "SESE")
  data = d
)
m2 <- lmer(
  height ~ scale(SDIinit) * factor(year) * Species + (1 | Site) + (1 | Plot) + (1 | Tree),
  # data = subset(d, Species == "SESE")
  data = d
)
m3 <- lmer(
  height ~ scale(SDIinit) * factor(year) * Species + (Species | Site) + (Species | Plot) + (Species | Tree),
  # data = subset(d, Species == "SESE")
  data = d
)

m
m2
m3

```


