---
title: Data import and description
---

<!--
source(knitr::purl("sprouts_import_data.qmd", tempfile()))
-->

```{r}
#| include: false

ggplot2::theme_update(
  panel.grid.major = ggplot2::element_blank(),
  panel.grid.minor = ggplot2::element_blank()
)
```

First, we'll load some libraries and our data.

```{r}
#| label: import-data

suppressWarnings(suppressPackageStartupMessages(library(tidyverse)))
library(ggridges)
library(ggokabeito)

sprouts <- readxl::read_excel("../data/Sprouts10yr.xlsx")

set.seed(743)
slice_sample(sprouts, n = 4) |> knitr::kable()
```

The data is in a wide format, variables are as follows:

```{r}
#| label: tbl-variable-descriptions
#| tbl-cap: Descriptions of variables in dataset.
#| echo: false

variable_descriptions <- list(
  Site = "One of four sites where treatments were replicated. Sites were located
  on similar slope positions, but across a range of aspects.",
  Plot = "There were 4 plots at each site and each was randomly assigned a
  treatment",
  Trtmt = "Treatemnt type: GS = group selection, which is basically a small
  clearing, LD = low density--fewer trees remaining; HD = high density--more
  trees remain and they are dispersed; HA = high density aggregated--more trees
  remain and they are grouped into clumps.",
  Tree = "Unique sprout ID.",
  Species = "SESE = coast redwood, and LIDE = tanoak.",
  HT1yr_m = "Sprout height one year after treatment.",
  HT5yr_m = "Same as above, but for year 5",
  HT10yr_m = "Same as above, but for year 10",
  `HT1-5yr` = "Height growth between years 1 and 5",
  `HT5-10yr` = "Height growth between years 5 and 10",
  DBH10yr_cm = "Diameter at breast height in cm at year 10. Only collected for
  redwood",
  LCBH10yr = "Live crown base height (height to first live branch) at year 10. 
  Only collected for redwood.",
  CR10yr = "Crown ratio (live crown length / total height) at year 10, only for
  redwood",
  HD10yr = "Unknown.",
  SDIinit = "Stand density index of plot immediately after treatment.",
  AggregatedYN = "Indicator for treatment HA.",
  ResidRWonClumpYN = "Unknown.",
  HT10rank = "Trees species specific height ranking within plot."
)

variable_descriptions |> 
  imap(\(x, y) tibble(Variable = y, Description = gsub("[[:space:]]{2,}", " ", x))) |>
  list_rbind(names_to = "Variable") |>
  knitr::kable()

```

Our main objectives are to:

- Test for treatment effects on redwood height and diameter at year 10
- Explore treatment effect on competition between redwood and tanoak
- Model the relationship between both of these and harvest intensity (SDIinit)
- Explore differences in height response over multiple remeasurements

The following figures reveal possible trends in the raw data.

@fig-data-summary-treatments shows that over time, LD and GS treatments have
more taller trees than the other treatments, with GS having the most. This is
true for both redwood and tanoak.

@fig-data-summary-spp seems to suggest that redwood are taller across all
treatments and years, but the GS treatment is the only one that provides a clear
advantage to redwood.

@fig-data-summary-sdi shows that above around 400 SDI, tree heights level off.
It also implies a steady decrease in height from 0 to around 400 SDI. The
strength of the relationship appears to be increasing over time, particularly
for redwood.

```{r}
#| label: fig-data-summary-treatments
#| fig-width: 8
#| fig-cap: >
#|   Height distributions for tanoak and redwood over time for each of four
#|   treatments. Treatments GS and LD have higher proportions of taller trees.
#|   Years refers to number of years after treatment.
#| message: false

# Function to convert heights from multiple years to long format
lengthen_heights <- function(data) {
  pivot_longer(data,
    matches("HT..?yr_m"),
    names_to = "year", values_to = "height",
    names_transform = \(x) as.integer(str_extract(x, "\\d+"))
  )
}

sprouts |>
  lengthen_heights() |>
  ggplot(aes(height, fct_rev(factor(year)), color = Trtmt)) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  geom_density_ridges(alpha = .3, size = 1) +
  facet_wrap(~Species) +
  labs(x = "Height (m)", y = "Year") +
  scale_color_okabe_ito()

```

```{r}
#| label: fig-data-summary-spp
#| fig-height: 8
#| fig-cap: >
#|   Similar to @fig-data-summary-treatments, but with an emphasis on
#|   differences between speceis responses across treatments. Comparisons are
#|   made for each year. The GS treatment appears to favor redwood response the
#|   most, but all treatments show redwoods are taller than tanoak. Years refers
#|   to number of years after treatment.
#| message: false

sprouts |>
  lengthen_heights() |>
  ggplot(aes(
    x = height,
    y = factor(Trtmt,levels = c("GS", "LD", "HA", "HD")),
    color = factor(Species, levels = c("SESE", "LIDE"))
  )) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  geom_density_ridges(alpha = .3, size = 1) +
  facet_grid(year ~ ., labeller = label_both) +
  labs(x = "Height (m)", y = "Treatment") +
  scale_color_okabe_ito(name = "Species")

```

```{r}
#| label: fig-data-summary-sdi
#| fig-cap: SDI vs HT for both species, across measurement years.
#| fig-width: 8.5

sprouts |>
  lengthen_heights() |>
  ggplot(aes(SDIinit, height, color = factor(year))) +
  geom_point(alpha = 0.5) +
  geom_smooth(span = 1.2) +
  facet_wrap(~Species) +
  scale_okabe_ito("color")

```

Our data though are nested, we have:

```
Sites
└─ Plots
   └─ Trees
      └─ Observations
```

Each Treatment is represented by one site/plot combination. Each site belongs to
each treatment and vice versa. I think the terminology here is that Sites and
treatments are crossed.


