---
title: Data import and description
format:
  html:
    code-fold: true
---

<!--
source(knitr::purl("sprouts_import_data.qmd", tempfile()))
-->

```{r}
#| include: false

# Quickly iterate over random color selections and pick the colors I like
rnd_color_brewer <- function(palette, chosen_order = NULL) {
  n <- RColorBrewer::brewer.pal.info[palette, "maxcolors"]
  ord <- sample(1:n)
  if (!is.null(chosen_order)) ord <- chosen_order
  colors <- RColorBrewer::brewer.pal(n, palette)[ord]
  print(ord)
  colors
}

ggplot2::theme_update(
  panel.grid.major = ggplot2::element_blank(),
  panel.grid.minor = ggplot2::element_blank()
)

```

## Getting started

First, we'll load some libraries and our data.

```{r}
#| output: false

library(lme4)
library(modelsummary)
library(marginaleffects)
library(ggdist)
library(ggridges)
library(emmeans)
library(multcomp)
library(multcompView)
library(tidyverse)
library(mgcv)
library(patchwork)
library(DHARMa)
library(glmmTMB)
library(ggeffects)
```

```{r}
#| echo: false

sprouts_fresh <- readxl::read_excel("../data/Sprouts10yr.xlsx")

set.seed(743)
slice_sample(sprouts_fresh, n = 4) |> knitr::kable()
```

The data is in a wide format, variables are as follows:

```{r}
#| label: tbl-variable-descriptions
#| tbl-cap: Descriptions of variables in dataset.
#| echo: false

variable_descriptions <- list(
  Site = "One of four sites where treatments were replicated. Sites were located
  on similar slope positions, but across a range of aspects.",
  Plot = "There were 4 plots at each site and each was randomly assigned a
  treatment",
  Trtmt = "Treatemnt type: GS = group selection, which is basically a small
  clearing, LD = low density--fewer trees remaining; HD = high density--more
  trees remain and they are dispersed; HA = high density aggregated--more trees
  remain and they are grouped into clumps.",
  Tree = "Unique sprout ID within Site, Plot, and Species",
  Species = "SESE = coast redwood, and LIDE = tanoak.",
  HT1yr_m = "Sprout height one year after treatment.",
  HT5yr_m = "Same as above, but for year 5",
  HT10yr_m = "Same as above, but for year 10",
  `HTI1-5yr` = "Height growth between years 1 and 5, (4 growth periods)",
  `HTI5-10yr` = "Height growth between years 5 and 10 (5 growth periods)",
  DBH10yr_cm = "Diameter at breast height in cm at year 10. Only collected for
  redwood",
  LCBH10yr = "Live crown base height (height to first live branch) at year 10. 
  Only collected for redwood.",
  CR10yr = "Crown ratio (live crown length / total height) at year 10, only for
  redwood",
  HD10yr = "Unknown.",
  SDIinit = "Stand density index of plot immediately after treatment.",
  AggregatedYN = "Indicator for treatment HA.",
  ResidRWonClumpYN = "Unknown.",
  HT10rank = "Trees species specific height ranking within plot."
)

variable_descriptions |>
  imap(function(x, y) {
    tibble(Variable = y, Description = gsub("[[:space:]]{2,}", " ", x))
  }) |>
  list_rbind(names_to = "Variable") |>
  knitr::kable()

```

## Wrangle

I'm going to change some variable names to make them more ergonomic

```{r}

newnames <- c(
  site = "Site", plot = "Plot", treat = "Trtmt", tree = "Tree", spp = "Species",
  ht1 = "HT1yr_m", ht5 = "HT5yr_m", ht10 = "HT10yr_m", ht_inc5 = "HTI1-5yr",
  ht_inc10 = "HTI5-10yr", dbh10 = "DBH10yr_cm", lcbh10 = "LCBH10yr",
  sdi_init = "SDIinit", agg = "AggregatedYN", ht_rnk10 = "HT10rank"
)

tibble(old = newnames, new = names(newnames)) |> knitr::kable()

```

Our data are nested. We have plots within sites, trees within plots, and
observeations within trees (multiple observations per tree).

```
Sites
└─ Plots
   └─ Trees
      └─ Observations
```

Each Treatment is represented by one site/plot combination. Each site belongs to
each treatment and vice versa. I think the terminology here is that Sites and
treatments are crossed.

Currently, `plot` (integer) is only unique within `site` and `tree` is only unique within
`site`, `treat`, and `spp`. It will be more convenient if `plot` and `tree` are
globally unique identifiers. This makes the nesting structure implicit and we
can simplify our model syntax.

```{r}
#| output: false

make_site_plot_unique <- function(data) {
  data |>
    mutate(plot = cur_group_id(), .by = c(site, treat)) |>
    mutate(tree = row_number())
}

```

I'm also going to order the treatments acording to our expectations about the
most to least productive. This will affect how they are plotted and reported.

```{r}

treat_order <- c("GS", "LD", "HA", "HD")

set_expected_treatment_order <- function(data) {
  data |>
    mutate(treat = fct_relevel(treat, treat_order))
}

```

For modeling purposes, I will make sure that the data are of the correct type.
Grouping variables and character data should be factors.

```{r}

# Assuming variable names have already been changed
set_data_types <- function(data) {
  data |>
    mutate(
      across(c(site, plot, treat, tree, spp), as.factor)
    )
}

```

And I'll apply all the above steps in one go.

```{r}

sprouts <- sprouts_fresh |>
  dplyr::select(all_of(newnames)) |>
  make_site_plot_unique() |>
  set_expected_treatment_order() |>
  set_data_types()

```

I will need a function to convert a variable in the data from wide to long format.

```{r}

lengthen_data <- function(data, var) {
  # I need regex for first part to handle ht and ht_inc and
  pref <- switch(var, ht = "(ht)", ht_inc = "(ht_inc)")
  # and the year
  suf <- "(\\d+)"
  str_to_match <- paste0(pref, suf)
  pivot_longer(data,
    matches(str_to_match),
    names_to = c(".value", "year"),
    names_pattern = str_to_match,
    names_transform = list(year = as.integer)
  ) |>
  relocate(year, matches(paste0(pref, "$")), .after = spp)
}

```

## Objectives

1. For each species/treatment combination, is there a difference in height
   increment between the first and second periods?
2. Is there a difference in redwood or tanoak sprout heights at year 10, if so,
   what is the magnitude of the difference?

## Visualize data

### Tree heights

The following figures reveal possible trends in the raw data.

#### Species/treatment/year

@fig-data-summary-treatments reveals fine, species specific differences between
treatments and general trends over time. It shows that over time, LD and GS
treatments have more taller trees than the other treatments, with GS having the
most. This is true for both redwood and tanoak. Additionally, for redwood it
appears that HA may have taller trees than HD, at least in year 10.

@fig-data-summary-spp shows the same information as before, but is arguably 
easier to look at. There is a general increasing trend in heights with
treatments as follows:

    HD < HA < LD < GS

Redwood is consistently taller than tanoak, and the GS treatment confers the
greatest advantage to redwood.

Across all treatments, it is also interesting to note that over times, the
height distributions, especially for redwood, seem to becoming more multi-modal
and more widely distributed. This could be due to site or plot effects, or to
microsite (within plot) effects, but it is not immediately clear why this
diverging performance should be so apparent with redwood and not tanoak. 


```{r}
#| label: fig-data-summary-treatments
#| fig-height: 8
#| fig-width: 8
#| fig-cap: >
#|   Height distributions for tanoak and redwood over time for each of four
#|   treatments. Treatments GS and LD have higher proportions of taller trees.
#|   Years refers to number of years after treatment.
#| message: false

sprouts |>
  lengthen_data("ht") |>
  ggplot(aes(ht, treat, color = treat, fill = treat)) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  geom_dots(binwidth = unit(0.019, "npc")) +
  facet_grid(vars(spp), vars(year), scales = "free", labeller = label_both) +
  labs(x = "Height (m)", y = "Treatment") +
  scale_color_brewer(palette = "Set2", aesthetics = c("color", "fill")) +
  theme(legend.position = "bottom")


```

```{r}
#| label: fig-data-summary-spp
#| fig-height: 8
#| fig-cap: >
#|   Similar to @fig-data-summary-treatments, but with an emphasis on
#|   differences between speceis responses across treatments. Comparisons are
#|   made for each year. The GS treatment appears to favor redwood response the
#|   most, but all treatments show redwoods are taller than tanoak. Years refers
#|   to number of years after treatment.
#| message: false

sprouts |>
  lengthen_data("ht") |>
  ggplot(aes(
    x = ht,
    y = fct_relevel(treat, treat_order),
    color = factor(spp, levels = c("SESE", "LIDE"))
  )) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  geom_density_ridges(alpha = .3, size = 1) +
  facet_grid(year ~ ., labeller = label_both) +
  labs(x = "Height (m)", y = "Treatment") +
  scale_color_brewer(palette = "Set2", name = "Species")
```

#### Species/year/SDI
@fig-data-summary-sdi shows that above around 400 SDI, tree heights level off.
It also implies a steady decrease in height from 0 to around 400 SDI. The
strength of the relationship appears to be increasing over time, particularly
for redwood.

```{r}
#| label: fig-data-summary-sdi
#| fig-cap: SDI vs HT for both species, across measurement years.
#| fig-width: 8.5

sprouts |>
  lengthen_data("ht") |>
  ggplot(aes(sdi_init, ht, color = factor(year))) +
  geom_point(alpha = 0.5) +
  geom_smooth(span = 1.2) +
  facet_wrap(~spp) +
  scale_color_brewer(palette = "Set2")
```

#### Species/site/plots
Lets see what the variability among sites and plots looks like, I'll focus on
year 10 only.

@fig-data-summary-sites reveals some differences between sites, particularly for
redwoods. Waldo North tends to have larger redwoods and Camp 6 has a large
proportion of smaller redwoods.

@fig-data-summary-plots shows that much of the difference in sites has to do
with one outlier plot within a site, and less about general site trends. We
should expect plots to capture a portion of the variance. Most notable is the
plot level difference between redwood and tanoak. For redwood, the large ammount
of within plot variability combined with the between plot variability obscures
the treatment (and site) effect. If you squint, there appears to be a similar
overall pattern between redwood and tanoak repsonse to treatment, but it appears
they respond differentially to certain plots.

```{r}
#| label: fig-data-summary-sites
#| fig-cap: >
#|   Distribution of heights at year 10 at each site for two species. Slabs are
#|   normalized by sample size to reflect the raw data, plotted as dots below.


sprouts |>
  ggplot(aes(ht10, site, fill = spp)) +
  stat_slab(aes(thickness = after_stat(pdf * n)), alpha = 0.6, scale = 0.7) +
  stat_dotsinterval(side = "bottom", scale = 0.7, slab_color = NA) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "year 10 height (m)", y = NULL, fill = "species")

```

```{r}
#| label: fig-data-summary-plots
#| fig-height: 7
#| fig-cap: >
#|   Same as for @fig-data-summary-sites, but for each plot (Site/treatment
#|   interaction). Grouped by treatment.

sprouts |>
  mutate(treat = fct_relevel(treat, treat_order)) |>
  ggplot(aes(
    ht10,
    site,
    fill = treat,
  )) +
  stat_slab(aes(thickness = after_stat(pdf * n)), alpha = 0.5, scale = 0.7) +
  stat_pointinterval(position = position_dodge(width = 0.5, preserve = "single")) +
  scale_color_brewer(palette = "Set2") +
  facet_wrap(~spp) +
  theme(legend.position = "bottom") +
  labs(y = NULL, x = "year 10 height (m)", fill = "treatment")

```

### Height increments

Height increments contain similar information as heights, but allow us to
compare directly between years.

@fig-data-summary-ht-inc-year shows that across treatments and species, height
growth slows down in the second period (years 5-10). This is more true for
redwood but it starts with more rapid growth than tanoak. In the most crowded
treatment (HD), redwoods height increment has become slower than tanoaks in the
second period. Also, in the second period, the high density, aggregated
treatment appears to have slightly higher (or equal) average growth increment,
which is not completely expected.

```{r}
#| label: fig-data-summary-ht-inc-year
#| fig-height: 7
#| fig-cap: >
#|   Variations in annual height growth increment between the first and second
#|   measurement periods (years 1-5, and 5-10, respectively).

sprouts |>
  lengthen_data("ht_inc") |>
  ggplot(
    aes(
      ht_inc,
      fct_relevel(treat, treat_order),
      fill = factor(year),
      color = factor(year)
    )
  ) +
  stat_slab(alpha = 0.5) +
  stat_pointinterval(
    position = position_dodge(width = 0.4, preserve = "single")
  ) +
  facet_wrap(~spp) +
  theme(legend.position = "bottom") +
  labs(
    fill = "Year",
    color = "Year",
    y = "Treatment",
    x = expression(Height~increment~(m~yr^-1))
  ) +
  scale_color_manual(
    values = rnd_color_brewer("Set2", c(1,4)),
    aesthetics = c("color", "fill")
  )

```

## Modeling

We have nested data where we have multiple observations within trees (over
time), trees within plots and plots within sites. Any of these levels of nesting
might result in correlations between observations, which violates the assumption
of independence for OLS models, but can be accounted for using multi-level
models.

I may want to include a categorical variable (spp) as a random slope (on the LHS
of the random parts) to allow random effects to be estimated for each species
separately. Michael Clark [explores this scenario], with some simpler data.

[explores this scenario]: https://m-clark.github.io/posts/2020-03-01-random-categorical

My approach will be to start with a simple model and build towards more complex
models.

### Height increment
By starting with height increment modeling, I will be able to answer question 1
in the objectives. This conclusion may be helpful for informing total height
modeling. I am curious whether height in the first 10 years increases linearly
with age. Is there a growth slow down in year 5-10 compared to years 1-5?

We saw in @fig-data-summary-ht-inc-year that increment does slow down somewhat
in the second measurement period (year 5-10), at least for redwoods, but we may
be justified in overlooking this. Lets look at individual tree's height
trajectories.

```{r}
#| label: fig-linear-ht
#| fig-height: 7
#| fig-cap: >
#|   Data plot showing trajectory of individual tree heights across the three
#|   measurements. Height increase appears fairly linear with year.

dht <- lengthen_data(sprouts, "ht")

dht |> 
  ggplot(aes(year, ht, color = spp, group = tree)) +
  geom_line(alpha = 0.6) +
  facet_wrap(~treat * site) +
  scale_color_brewer(palette = "Set2") +
  theme(
    legend.position = "bottom"
  )

```

```{r}
#| echo: false
#| output: asis

dinc <- lengthen_data(sprouts, "ht_inc") |> mutate(year = as.factor(year))

minc <- lst(
  minc1 = lm(ht_inc ~ year, dinc),
  minc2 = update(minc1, ~ . * spp),
  minc3 = update(minc2, ~ . * treat),
  minc4 = update(minc1, ~ . + spp + treat),
  minc5 = update(minc4, ~ . + spp:treat),
  minc6 = update(minc5, ~ . + spp:year),
  minc7 = update(minc6, ~ . + treat:year),
  minc8 = lmer(
    ht_inc ~ treat + year + spp + treat:spp + spp:year + (1 | plot) +
      (1 | tree),
    dinc
  ),
  minc9 = update(minc8, ~ . + treat:year),
  minc10 = update(minc8, ~ . - (1 | plot) + (0 + spp | plot)),
  minc11 = update(minc10, ~ . - treat:spp),
)

format_model_formula <- function(mod) {
  form <- format(formula(mod)) |>
  paste0(collapse = "") |>
  str_replace("\\s{2,}", " ")
  form
}

label_model_formula <- function(mod, name) {
  form <- format_model_formula(mod)
  form <- paste0("`", form, "`")
  cat(paste(name, form, sep = ": "), "\n\n")
}

# iwalk(minc, label_model_formula)

```

For the most part, it looks like height increases linearly with year. Lets test
this with a model.

1. ``r format_model_formula(minc[[1]])``

   Ignoring everything else, there is a 0.08 m / yr slow down in the second
   period.

2. ``r format_model_formula(minc[[2]])``

   That slow down has more to do with redwood than tanoak.

3. ``r format_model_formula(minc[[3]])``
   
   There are overall treatment effect on growth which varies by species (redwood
   is more affected) but there are no more significant interactions with year.
   Our data does not detect significant differences in slow down by treatment or
   species.
   j
4. ``r format_model_formula(minc[[4]])``

   This includes only additive effects and the effect of year is the same as
   model 1.
   
5. ``r format_model_formula(minc[[5]])``

   Again, the species by treatment interactions seem to be significant.
   
6. ``r format_model_formula(minc[[6]])``

   The species year interaction remains important.

7. ``r format_model_formula(minc[[7]])``

   Again, the treatment by year interactions do not seem important/estimable.

8. ``r format_model_formula(minc[[8]])``

   *Now I introduce random intercepts for plot and tree.* These lead to *much*
   smaller AIC's.

9. ``r format_model_formula(minc[[9]])``

   But treatment by year is still (mostly) not significant.

10. ``r format_model_formula(minc[[10]])`` 

    Adding a random slope for species by plot results in another big jump in
    AIC, but also reduces confidence in the treatment*species interaction. It
    would seem that the potential differences in species by treatment growth
    rate are obscured by redwoods greater variability across plots.

11. ``r format_model_formula(minc[[11]])`` 
   
    Removing the treat * species interaction results in a solid AIC boost. 

```{r}
#| echo: false
#| column: body-outset

# lmer and lm name coefficients differently, leading to duplicates in the
# table
rearrange_coef_names <- function(x) {
  str_replace(x, "^(.*):(treat[HALD]{2})$", "\\2:\\1") |>
    str_replace(":", " x ")
}

modelsummary(
  minc,
  stars = TRUE,
  coef_rename = rearrange_coef_names,
  output = "markdown",
  metrics = c("AIC", "BIC", "RMSE"),
  estimator = "ML",
  gof_omit = "F|Log.Lik"
)

```

### Height increment results

The model selected based on AIC lacks a treatment x year interaction. This means
that there was not enough evidence to suggest that any treatment results in a
lesser or greater slow down in height growth. Across treatments, the effect of
year was significant, and this varied significantly by species--redwood's height
growth rate reduction was about 3 times that of tanoak.

```{r}
#| label: fig-ht-inc-contrast
#| fig-cap: >
#|   Growth period 1 is year 1-5, growth period 2 is year 5-10. Growth
#|   significantly decreased in the second period for both species.

minc11em <- emmeans(minc[[11]], c("year"), by = c("spp"))

minc11con <- contrast(minc11em, "consec") |> confint() |> as_tibble()

minc11cld <- cld(minc11em, Letters = letters) |>
  as_tibble() |>
  mutate(.group = str_replace_all(.group, " ", ""))

minc11con |>
  ggplot(aes(spp, -estimate, color = spp)) +
  geom_pointrange(
    aes(ymin = -lower.CL, ymax = -upper.CL),
    linewidth = 1, size = 1
    ) +
  scale_color_brewer(palette = "Set2", aesthetics = c("color", "fill")) +
  labs(
    x = NULL,
    y = expression(Height ~ increment ~ reduction ~ (-m %.% yr^-1)),
    color = "Species"
  )

```

```{r}
#| label: tbl-ht-inc-contrast
#| tbl-cap: >
#|   Growth period 1 is year 1-5, growth period 2 is year 5-10. Growth
#|   significantly decreased in the second period for both species.

minc11con |> knitr::kable(digits = 3)

```

```{r}
#| label: fig-ht-inc
#| fig-cap: >
#|   There was a significant difference in height increment between period 1 and
#|   2 for both species.

minc11cld |>
  ggplot(aes(year, emmean, fill = spp, color = spp)) +
  geom_pointrange(
    aes(ymin = lower.CL, ymax = upper.CL),
    position = position_nudge(x = -0.15),
    size = 0.9, linewidth = 1
  ) +
  geom_dots(data = dinc, aes(year, ht_inc), alpha = 0.4) +
  facet_grid(~spp, switch = "x") +
  theme(
    panel.spacing = unit(0, "lines"),
    strip.background = element_blank(),
    strip.text = element_blank(),
  ) +
  geom_text(
    aes(y = upper.CL, label = .group),
    # color = "black",
    position = position_nudge(x = -0.15, y = 0.1)
  ) +
  scale_color_brewer(palette = "Set2", aesthetics = c("color", "fill")) +
  labs(
    x = "Growth period",
    y = expression(Height ~ increment ~ (m %.% year^-1)),
    color = "Species",
    fill = "Species"
  ) +
  scale_x_discrete(labels = c("1", "2"))

```

### Heights - year 10 only
The simplest model that I'm willing to look at includes a treatment/species
interaction, and no random effects. It will serve as a baseline for comparison.


```{r}
#| code-fold: false

# I'm calling the height data "d"
dht10 <- sprouts |> lengthen_data("ht") |> filter(year == 10)

mht101 <- lm(ht ~ treat * spp, data = dht10)
mht102 <- lmer(ht ~ treat * spp + (1 | plot) + (1 | site), data = dht10)
mht103 <- lmer(ht ~ treat * spp + (1 | plot), data = dht10)
mht104 <- lmer(ht ~ treat * spp + (0 + spp | plot), data = dht10)
mht105 <- glmmTMB(
  ht ~ treat * spp + (0 + spp | plot),
  family = Gamma(link = "log"),
  data = dht10
)
mht106 <- glmmTMB(
  ht ~ treat * spp + (0 + spp | plot),
  family = Gamma(link = "log"),
  data = dht10,
  dispformula = ~ spp
)
mht107 <- glmmTMB(
  ht ~ treat + spp + (0 + spp | plot),
  family = Gamma(link = "log"),
  data = dht10,
  dispformula = ~ spp
)
mht108 <- glmmTMB(
  ht ~ treat + spp + (0 + spp | plot),
  family = Gamma(link = "log"),
  data = dht10,
  dispformula = ~ spp + treat
)

mht10_list <- lst(
  mht101, mht102, mht103, mht104, mht105, mht106, mht107, mht108
)

modelsummary(
  mht10_list,
  stars = TRUE,
  output = "markdown",
  shape = term:component + statistic ~ model,
  metrics = c("AIC", "BIC", "RMSE"),
  estimator = "ML",
  gof_omit = "F|Log.Lik"
)

```
The best model according to AIC includes a random intercept for plot and a
random slope for spp. Because spp is a factor, this effectively estimates a
species specific intercept, but also takes into account the covariance of the
species plot effect. The best model for dispersion included speceis and
treatment.

### Year 10 only Results

```{r}
#| label: fig-year-10-only
#| fig-cap: >
#|   Mean and 95% confidence interval for the selected model, displayed with raw
#|   data. Shared letters between treatments (within species) indicates that
#|   there is not sufficient evidence to claim the estiamtes come from different
#|   distributions.

mht108grid <- emmeans(mht108, pairwise ~ treat, by = c("spp"), type = "response")

mht108cld <- cld(mht108grid, Letters = letters) |>
  as_tibble() |>
  mutate( .group = str_replace_all(.group, " ", ""))

year_10_only_plot <- dht10 |>
  ggplot(aes(treat, ht, color = spp, fill = spp)) +
  facet_grid(~spp, switch = "x") +
  theme(panel.spacing = unit(0, "lines"),
        strip.background = element_blank(),
        strip.text = element_blank(),
        ) +
  geom_dots() +
  scale_color_brewer(palette = "Set2", aesthetics = c("color", "fill")) +
  geom_pointrange(
    data = mht108cld,
    aes(y = response, ymin = asymp.LCL, ymax = asymp.UCL),
    color = "gray50",
    position = position_nudge(x = -0.07),
    size = 0.7,
    linewidth = 1,
    show.legend = FALSE
  ) +
  geom_text(
    data = mht108cld,
    aes(y = asymp.UCL, label = .group),
    color = "black",
    position = position_nudge(x = -0.3, y = 0.2)
  )

year_10_only_plot

```

### Heights - multiple years

By including year as a numeric variable, we are assuming a liner relationship
between year and height. This term (and it's interactions) can than be
interpreted as a modeled growth increment, and although we only have
observations at years 1, 5, and 10, other years predictions will be linearly
interpolated.

The following models were fit, and are summarized in @tbl-model-summaries.

2. OLS linear model including year in a 3-way interaction.
3. Eliminate the 3-way interaction (`spp:treat:year`), keeping others. This
   results in a significantly lower AIC.
4. First random-effects model including interactions and universal random slopes
   for site and plot, and time-series.
5. Allow the random effects to (co-) vary by species.
6. Same as 5, but attempt to remove the 3-way interaction again.
7. Same as 6, but attempt to simplify the random effects by forcing them to be
   estimated independently, correlation between species is not estimated.
8. Re-introduce the 3-way interaction again.
9. Omit the random effect for site. This is supported by the interpretation of @fig-data-summary-plots.
10. Like 9, but try simplified (independent) random effects.

One problem with this is that I'm comparing AIC for different random effects
using ML, when I should be using REML. This may be a problem throughout my model
selection.

```{r}
#| code-fold: false
#| label: tbl-model-summaries
#| tbl-cap: >
#|   Summary table for models 2-10, described above. Standard error in
#|   parantheses.
#| column: body-outset

dht <- lengthen_data(sprouts, "ht")

mht2 <- lm(ht ~ treat * spp * year, data = dht)

mht3 <- lm(ht ~ treat + spp + year + treat:year + spp:year + treat:spp, data = dht)

mht4 <- lmer(
  ht ~ treat * spp * year + (1 | site) + (1 | plot) + (1 | tree),
  data = dht
)

mht5 <- lmer(
  ht ~ treat * spp * year + (0 + spp | site) + (0 + spp | plot) + 
    (0 + spp | tree),
  data = dht
)

mht6 <- lmer(
  ht ~ treat + spp + year + treat:spp + year:spp + treat:year +
    (0 + spp | site) + (0 + spp | plot) + (0 + spp | tree),
  data = dht
)

mht7 <- lmer(
  ht ~ treat + spp + year + treat:spp + year:spp + treat:year + (1 | site:spp) +
    (1 | plot:spp) + (1 | tree:spp),
  data = dht
)

mht8 <- lmer(
  ht ~ treat * spp * year + (1 | site:spp) + (1 | plot:spp) + (1 | tree:spp),
  data = dht
)

mht9 <- lmer(
  ht ~ treat * spp * year + (0 + spp | plot) + (0 + spp | tree),
  data = dht
)

mht10 <- lmer(
  ht ~ treat * spp * year + (1 | plot:spp) + (1 | tree:spp),
  data = dht
)

# I tried fitting year as a factor, but then could not fit a model that included
# the 3-way interaction and random slopes for species. The model without the
# 3-way (and spp random slopes) was close, but about 12 AIC larger.

modelsummary(
  list(
    "Model 2" = mht2, "Model 3" = mht3, "Model 4" = mht4, "Model 5" = mht5,
    "Model 6" = mht6, "Model 7" = mht7, "Model 8" = mht8, "Model 9" = mht9,
    "Model 10" = mht10
  ),
  stars = TRUE,
  output = "markdown",
  metrics = c("AIC", "BIC", "RMSE"),
  estimator = "ML",
  gof_omit = "F|Log.Lik"
)

```

### Multi-year heights V2

The problem with all of the above models is (a) a normal distribution is not a
good fit for the data because it results in negative height predictions for
small trees (year 1) and (b) variance is not constant. Finally, we are fitting
year as a linear effect because fitting it as a factor with 3 levels led to
model convergence problems, but ideally, we would like to allow different
relationships between height and year, because, as we saw in the previous
section, height growth is not constant in the first 10 years.

For these reasons it makes sense to try another modeling framework,
particularly, a gamma distributed model, with a modeled dispersion parameter. It
allows only positive predictions, can right skewed distributions, and the
dispersion can be modeled as a function of covariates.

```{r}
#| code-fold: false
#| label: tbl-model-summaries-v2
#| tbl-cap: >
#|   Summary table for models 2-10, described above. Standard error in
#|   parantheses.
#| column: body-outset

mht11 <- glmmTMB(
  ht ~ treat * spp * year + (1 | plot) + (1 | tree),
  data = dht, family = Gamma(link = "log")
)

mht12 <- glmmTMB(
  ht ~ treat + spp + year + treat:spp + treat:year + spp:year + (1 | plot) + 
  (1 | tree),
  data = dht, family = Gamma(link = "log")
)

mht13 <- glmmTMB(
  ht ~ treat + spp + year + treat:spp + treat:year + spp:year + 
  (0 + spp | plot) + (0 + spp | tree),
  data = dht, family = Gamma(link = "log")
)

mht14 <- glmmTMB(
  ht ~ treat + spp + year + treat:spp + treat:year + spp:year +
  (1 | plot) + (1 | tree),
  data = mutate(dht, year = factor(year)), family = Gamma(link = "log")
)

mht15 <- glmmTMB(
  ht ~ treat + spp + year + treat:spp + treat:year + spp:year +
  (1 | plot) + (1 | tree),
  dispformula = ~ year,
  data = mutate(dht, year = factor(year)), family = Gamma(link = "log")
)

mht16 <- glmmTMB(
  ht ~ treat + spp + year + treat:spp + treat:year + spp:year +
  (1 | plot) + (1 | tree),
  dispformula = ~ year * spp,
  data = mutate(dht, year = factor(year)),
  family = Gamma(link = "log")
)

mht17 <- glmmTMB(
  ht ~ treat + spp + year + treat:spp + treat:year + spp:year +
  (1 | plot) + (1 | tree),
  dispformula = ~ year * spp + treat + treat:spp,
  data = mutate(dht, year = factor(year)), family = Gamma(link = "log")
)

mht18 <- glmmTMB(
  ht ~ treat + spp + year + treat:year + spp:year +
  (1 | plot) + (1 | tree),
  dispformula = ~ year + spp + treat + treat:spp,
  data = mutate(dht, year = factor(year)),
  family = gaussian(link = "log")
)

do_model_summary_table <- function(model_list) {
  modelsummary(
    model_list,
    output = "markdown",
    shape = term:component + statistic ~ model,
    metrics = c("AIC", "BIC", "RMSE"),
    estimator = "ML",
    gof_omit = "F|Log.Lik",
    stars = TRUE
  )
}

do_model_summary_table(list(
    "Model 11" = mht11, "Model 12" = mht12, "Model 13" = mht13,
    "Model 14" = mht14, "Model 15" = mht15, "Model 16" = mht16,
    "Model 17" = mht17, "Model 18" = mht18
))

do_model_summary_table(lst(mht18, mht19))

lst(mht18, mht19)

```

The model with the lowest AIC is model 18.

### Multi-year model checking

```{r}
#| label: fig-multi-year-resid
#| fig-cap: >
#|   Residual vs fitted plot for the selected multi-year linear regression model
#|   for heights.

res <- simulateResiduals(mht18)
plot(res)

```

### Multi-year results

Using model 18, we can now answer the questions we posed in the Objectives
section.

```{r}
#| label: fig-multi-year-growth
#| fig-height: 7
#| fig-cap: >
#|   Average species and treatment specific height as a linear function
#|   of year. Density distributions represent the raw data associated with the
#|   estimate at that point. The grey line is the OLS fit for reference.

mht18emmeans <- emmeans(
  mht18,
  c("spp", "treat", "year"),
  type = "response"
)

mht18emmeans |>
  as_tibble() |>
  mutate(
    year = as.numeric(levels(year))[year]
  ) |>
  rename(any_of(c(lower.CL = "asymp.LCL", upper.CL = "asymp.UCL"))) |>
  ggplot(aes(year, response, color = spp, fill = spp, group = spp)) +
    geom_line() +
    facet_wrap(~ treat) +
    geom_ribbon(
      aes(ymin = lower.CL, ymax = upper.CL, color = NULL),
      alpha = .2
    ) +
    stat_slab(data = dht, aes(year, ht), alpha = 0.4) +
    stat_smooth(
      data = dht,
      aes(year, ht),
      method = lm,
      color = "gray60",
      se = FALSE,
      linewidth = 0.7
    ) +
    scale_color_brewer(palette = "Set2", aesthetics = c("color", "fill")) +
    theme(legend.position = "bottom") +
    labs(x = "Year", y = "Height (m)", color = "Species", fill = "Species")

```

```{r}
#| label: fig-multi-year-at-10
#| fig-cap: >
#|   Marginal means at year 10 based on height data collected three times over
#|   10 years. The same letter within a species indicates that there was not
#|   enough evidence to differentiate those treatments.

mht18_compare <- emmeans(
  mht18,
  spec = "treat",
  by = c("spp"),
  at = list(year = "10"),
  type = "response"
) |>
  cld(Letters = letters) |>
  as_tibble() |>
  mutate(
    # year = as.numeric(levels(year))[year],
    .group = str_replace_all(.group, " ", "")
  ) |>
  rename(any_of(c(lower.CL = "asymp.LCL", upper.CL = "asymp.UCL")))

multiple_years_plot <- filter(dht, year == 10) |>
  ggplot(aes(treat, ht, color = spp, fill = spp)) +
  facet_grid(~spp, switch = "x") +
  theme(panel.spacing = unit(0, "lines"),
        strip.background = element_blank(),
        strip.text = element_blank(),
        ) +
  geom_dots() +
  scale_color_brewer(palette = "Set2", aesthetics = c("color", "fill")) +
  geom_pointrange(
    aes(y = response, ymin = lower.CL, ymax = upper.CL),
    data = mht18_compare,
    color = "gray50", position = position_nudge(x = -0.07),
    size = 0.7,
    linewidth = 1,
    show.legend = FALSE
  ) +
  geom_text(
    aes(y = upper.CL, label = .group),
    data = mht18_compare,
    color = "black",
    position = position_nudge(x = -0.3, y = 0.2)
  ) +
  labs(x = "Treatment", y = "Height (m)", color = "Species", fill = "Species") +
  theme(legend.position = "bottom")

multiple_years_plot

```


### Multi-year vs. year 10 only

```{r}
standard_model_col_names <- c(
  LCL = "lower.CL",  UCL = "upper.CL",  LCL = "asymp.LCL", UCL = "asymp.UCL",
  emmean = "response"
)

list("year 10 only" = mht108cld, "multi-year" = mht18_compare) |>
  map(\(x) rename(x, any_of(standard_model_col_names))) |>
  bind_rows(.id = "model") |>
  ggplot(aes(treat, ht, color = model, fill = model)) +
  facet_grid(~spp) +
  geom_pointrange(
    aes(y = emmean, ymin = LCL, ymax = UCL),
    position = position_dodge(width = 0.4),
    size = 0.7,
    linewidth = 1
  ) +
  geom_dots(
    data = dht10, color = "gray50", fill = "gray70",
    position = position_nudge(x = 0.2), scale = 0.7, binwidth = 0.18
  ) +
  geom_text(
    aes(y = UCL + 0.6, label = .group, color = model,),
    position = position_dodge(width = 0.4),
  ) +
  theme(legend.position = "bottom") + 
  scale_color_manual(
    values = rnd_color_brewer("Set2", c(1,4)),
    aesthetics = c("color", "fill")
  )

```
